

\begin{center}\begin{large} Practice Problems 4
 \end{large}\end{center}
 \bigskip


\tableofcontents 
 
% 1 - 2 Children 
\begin{problem} (2 children)
    Consider a family that has two children. We are interested in the children's genders. Our sample space is:
    \[
    S = \{(G, G), (G, B), (B, G), (B, B)\}
    \]
    where \( G \) represents a girl and \( B \) represents a boy. Assume that all four possible outcomes are equally likely.
    
    \begin{enumerate}
        \item What is the probability that both children are girls given that the first child is a girl?
        
        \item Suppose we ask the father: "Do you have at least one daughter?" He responds "Yes!" Given this extra information, what is the probability that both children are girls? In other words, what is the probability that both children are girls given that we know at least one of them is a girl?
    \end{enumerate}
    \href{https://www.probabilitycourse.com/chapter1/1_4_0_conditional_probability.php#example18}{Solution (Ex. 1.18)}

\end{problem}  




% 2 - 2 Children, but we know the name
\begin{problem} (2 Children, one Lilia) 
    A family has two children. We ask the father, "Do you have at least one daughter named Lilia?" He replies, "Yes!" What is the probability that both children are girls? 

    In other words, we want to find the probability that both children are girls, given that the family has at least one daughter named Lilia. Here, you can assume the following:
    \begin{itemize}
        \item If a child is a girl, her name will be Lilia with probability \( \alpha \ll 1 \), independently from other children's names.
        \item If the child is a boy, his name will not be Lilia.
    \end{itemize}
\end{problem}
\href{https://www.probabilitycourse.com/chapter1/1_4_5_solved3.php}{Solution (Pr. 7)}

% 3 - Elections
\begin{problem} (elections)
In a certain town, 30\% of the people are conservatives and 70\% socialists. At the last election, 65\% of conservatives voted and 80\% of socialists. A person from the town is selected at random, and states that she voted at the last election. What is
the probability that she is a socialist?
\end{problem}

% 4 - Spam emails 
\begin{problem} (Spam emails)
        A spam filter tags emails as spam or not spam. Based on historical data:
    \begin{itemize}
        \item 80\% of spam emails contain the word "lottery."
        \item 30\% of non-spam emails contain the word "lottery."
        \item 40\% of emails are spam.
    \end{itemize}
    
    What is the probability that an email is spam given it contains the word "lottery"?
\end{problem}
% 5 - PMF, CDF, 2 coins
\begin{problem} (PMF and CDF for 2 coin flips)
    I toss a coin twice. Let $X$  be the number of observed heads. Find and plot the PMF and CDF of $X$.

\href{https://www.probabilitycourse.com/chapter3/3_1_3_pmf.php}{Solution for the PMF (Ex. 3.3)}, \href{https://www.probabilitycourse.com/chapter3/3_2_1_cdf.php}{Solution for the CDF (Ex 3.9)}
\end{problem}
% 6 - PDF, CDF
\begin{problem} (PDF, CDF)
    Let \( X \) be a continuous random variable with the following PDF:
    \[
    f_X(x) =
    \begin{cases} 
    c e^{-x} & \text{if } x \geq 0, \\
    0 & \text{otherwise,}
    \end{cases}
    \]
    where \( c \) is a positive constant.
    
    \begin{enumerate}
        \item Find \( c \).
        \item Find the CDF of \( X \), denoted \( F_X(x) \).
        \item Find \( P(1 < X < 3) \).
    \end{enumerate}
\href{https://www.probabilitycourse.com/chapter4/4_1_1_pdf.php}{Solution (Ex. 4.2)}
\end{problem}


% 7 - Expected value and Variance of the Uniform
\begin{problem} (Expected value and Variance of the Uniform distribution)
Let \( X \sim \text{Uniform}(a, b) \). 

\begin{enumerate}
    \item Find \( \mathbb{E}[X] \) (the expected value of \( X \)).
    \item Find \( \mathbb{V}\text{ar}(X) \) (the variance of \( X \)).
\end{enumerate}

\href{https://www.probabilitycourse.com/chapter4/4_1_2_expected_val_variance.php}{Solution for 1. (Ex 4.3)}
\end{problem}

    

% \[
% \text{Var}(X) = E\left[(X - \mu_X)^2\right] = E[X^2] - (E[X])^2.
% \]