\documentclass[aspectratio=169]{beamer}
\input{../preamble}

\usepgfplotslibrary{groupplots, fillbetween}
\usepackage{listings}

\lstset{
  basicstyle=\ttfamily\footnotesize,
  keywordstyle=\color{armblue}\bfseries,
  commentstyle=\color{paramgreen}\itshape,
  stringstyle=\color{armred},
  showstringspaces=false,
  language=Python,
  frame=single,
  framerule=0pt,
  backgroundcolor=\color{lightbg},
  numbers=none,
  breaklines=true,
  xleftmargin=4pt,
  xrightmargin=4pt,
}

\title{L01e: Hyperparameter Tuning}
\subtitle{Manual, Grid, Random, and Bayesian Optimization}
\date{}

\begin{document}

\begin{frame}{Outline}
  \tableofcontents
\end{frame}

% ================================================================
\section{Why tune?}

\begin{frame}{Parameters vs.\ hyperparameters}
  \textbf{The simple test:} can you \emph{compute} it from the data alone, or do you have to \emph{choose} it? Compute $\Rightarrow$ parameter. Choose $\Rightarrow$ hyperparameter.
  \vfill
  \begin{columns}[T]
    \column{0.5\textwidth}
    \textbf{Parameters $\thetav$} are what the model \emph{learns} from data.
    \begin{itemize}\small
      \item Linear regression: the coefficients (one per feature + intercept).
      \item Decision tree: the split points and feature choices at each node.
      \item Neural net: all the weights and biases.
    \end{itemize}
    \column{0.5\textwidth}
    \textbf{Hyperparameters $\lamv$} are what the model \emph{doesn't} learn - you have to choose them.
    \begin{itemize}\small
      \item Ridge regression: \texttt{alpha} (regularization strength).
      \item Decision tree: \texttt{max\_depth}, \texttt{min\_samples\_leaf}.
      \item Neural net: learning rate, number of layers, batch size, weight decay.
    \end{itemize}
  \end{columns}
  \vfill
  Different hyperparameter choices $\Rightarrow$ different fitted models $\Rightarrow$ different generalization. \textbf{Tuning} = systematically picking $\lamv$ to optimize a validation/CV score.
\end{frame}

\begin{frame}{Predict-first: how many hyperparameters?}
  \pause
  \textbf{Predict the count of meaningfully tunable hyperparameters for each model:}
  \begin{itemize}
    \item Ridge regression (sklearn)
    \item Random forest (sklearn)
    \item XGBoost
    \item GPT-style transformer training
  \end{itemize}
  \vfill
  \pause
  Rough answers:
  \begin{itemize}
    \item \textbf{Ridge:} 1 (\texttt{alpha}). Most people only tune this.
    \item \textbf{Random forest:} 4-6 (\texttt{n\_estimators}, \texttt{max\_depth}, \texttt{min\_samples\_leaf}, \texttt{max\_features}, \texttt{bootstrap}).
    \item \textbf{XGBoost:} 10-15 sensitive ones (\texttt{learning\_rate}, \texttt{max\_depth}, \texttt{subsample}, \texttt{colsample\_bytree}, \texttt{min\_child\_weight}, regularization, $\ldots$).
    \item \textbf{Transformer training:} dozens (architecture + optimization + data + regularization).
  \end{itemize}
  Tuning effort scales with the search-space size. \textbf{The number of hyperparameters dictates which search method makes sense:} Ridge ($1$ HP) $\to$ manual or grid; XGBoost ($15$) $\to$ random or Bayesian. The next sections develop this map.
\end{frame}

\begin{frame}{The bilevel optimization view}
  Tuning is an \emph{outer} optimization wrapping the model's own \emph{inner} optimization.
  \vfill
  \begin{center}
    \begin{tikzpicture}[
      every node/.style={font=\small},
      outer/.style={rectangle, draw=armblue, very thick, rounded corners=4pt, fill=armblue!8, inner sep=8pt, minimum width=11.5cm, minimum height=2.6cm},
      inner/.style={rectangle, draw=paramgreen, very thick, rounded corners=4pt, fill=paramgreen!10, inner sep=6pt},
    ]
      \node[outer] (outer) {};
      \node[anchor=north west, font=\small\bfseries, armblue] at (outer.north west) {\;Outer loop (tuning): pick best $\lamv^*$};
      \node[inner, below=0.5cm of outer.north, anchor=north] (inner) {
        \begin{minipage}{7.5cm}\centering
          \textbf{Inner loop (training):} given $\lamv$, solve
          $\thetah(\lamv) = \argmin_{\thetav}\, \riske(\thetav;\, \lamv)$
        \end{minipage}};
      \node[anchor=south, font=\scriptsize, armblue!80!black] at ([yshift=0.05cm]outer.south) {pass $\lamv$ in $\to$ get back $\thetah(\lamv)$, score it on validation/CV $\to$ outer search picks next $\lamv$};
    \end{tikzpicture}
  \end{center}
  \vfill
  \footnotesize \textbf{In plain English:} the outer loop tries hyperparameter values; the inner loop fits the model with those values. \emph{One outer step = one full model training.}
  \vskip 4pt
  Choices on outer search strategy:
  \begin{itemize}\small
    \item \textbf{Manual} - you propose $\lamv$ values from intuition.
    \item \textbf{Grid} - exhaustive over a Cartesian product.
    \item \textbf{Random} - sample uniformly from distributions.
    \item \textbf{Bayesian} - build a surrogate model of score vs.\ $\lamv$, query smartly.
  \end{itemize}
\end{frame}

% ================================================================
\section{Manual tuning}

\begin{frame}{Manual tuning: ``grad student descent''}
  The simplest strategy: a human proposes one $\lamv$, evaluates, then proposes another based on what they saw.
  \vfill
  Typical workflow:
  \begin{enumerate}
    \item Start with the library's defaults. Get a baseline.
    \item Identify the most-likely-impactful hyperparameter (domain knowledge or quick experiments).
    \item Sweep it across $\sim 5$ values, plot validation score, pick the best.
    \item Fix that value, move to the next hyperparameter, repeat.
    \item Stop when changes don't improve validation score.
  \end{enumerate}
  \vfill
  Also called \emph{manual coordinate descent} - you adjust one hyperparameter at a time.
  \vfill
  \fcolorbox{armred}{armred!8}{\parbox{0.96\textwidth}{\small
    \textbf{Manual is fast to start and slow to scale.} It works for 1-2 HPs and an experienced practitioner. For everything else, the automated methods coming up are usually better. Treat the rest of this deck as upgrades, not replacements.
  }}
\end{frame}

\begin{frame}{When manual works (and when it doesn't)}
  \begin{columns}[T]
    \column{0.5\textwidth}
    \textbf{Manual is best when:}
    \begin{itemize}\small
      \item You have strong intuition about the model (you've trained 50 Ridge regressions already).
      \item The search space is small (1-2 hyperparameters).
      \item Each evaluation is slow (you don't want to launch 100 jobs by accident).
      \item You're \emph{developing} a sense for the model, not just optimizing.
    \end{itemize}
    \column{0.5\textwidth}
    \textbf{Manual breaks down when:}
    \begin{itemize}\small
      \item More than $\sim 3$ hyperparameters interact (you can't visualize the surface).
      \item You can't afford to be inconsistent across runs.
      \item Multiple people work on the same model (reproducibility).
      \item You're shipping to production and need a defensible methodology.
    \end{itemize}
  \end{columns}
  \vfill
  Manual tuning teaches you the most per evaluation but scales the worst.
\end{frame}

\begin{frame}{Pitfalls of manual tuning}
  \begin{enumerate}
    \item \textbf{Confirmation bias.} You notice the run that worked, forget the three that didn't. Mitigation: write down every run, including failures.
    \item \textbf{Validation-set abuse.} The 50th hyperparameter setting you tried still gets evaluated on the same validation set. You're effectively overfitting validation. Mitigation: cross-validation, separate test set, count your evaluations.
    \item \textbf{No reproducibility.} ``I'll try \texttt{alpha=0.3} next... or was it 0.03?'' Mitigation: a notebook log, or skip manual and use a configured automated search.
    \item \textbf{Local optimum.} Coordinate descent gets stuck. The best Ridge \texttt{alpha} \emph{given the current data preprocessing} may differ from the best after you change features. Mitigation: redo the whole sweep occasionally.
  \end{enumerate}
\end{frame}

% ================================================================
\section{Grid search}

\begin{frame}{Grid search: exhaustive over a Cartesian product}
  Pick a finite list of candidate values for each hyperparameter. Evaluate \emph{every} combination.
  \vfill
  Example for Ridge with polynomial features:
  \begin{center}\small
    \begin{tabular}{ll}
      Hyperparameter & Candidate values \\
      \hline
      \texttt{alpha} & $\{10^{-3}, 10^{-2}, 10^{-1}, 1, 10, 100\}$ \\
      \texttt{poly\_degree} & $\{1, 2, 3, 4, 5\}$ \\
    \end{tabular}
  \end{center}
  \vfill
  Total: $6 \times 5 = 30$ configurations. With 5-fold CV, $30 \times 5 = 150$ model fits.
  \vfill
  Exhaustive, reproducible, embarrassingly parallel. Also potentially \emph{very} expensive.
\end{frame}

\begin{frame}{Predict-first: combinatorial blowup}
  You want to tune XGBoost with 4 hyperparameters, 5 candidate values each, using 5-fold CV.
  \vfill
  \pause
  \textbf{Predict:}
  \begin{itemize}
    \item How many model fits will grid search require?
    \item If each fit takes 30 seconds, how long does the whole search take (single-threaded)?
  \end{itemize}
  \vfill
  \pause
  Answers:
  \begin{itemize}
    \item $5^4 \times 5 = 625 \times 5 = \mathbf{3125}$ model fits.
    \item $3125 \times 30\text{s} \approx 26$ hours single-threaded.
  \end{itemize}
  \vfill
  Add one more hyperparameter (5 values): $\mathbf{15{,}625}$ fits, $\mathbf{5.4}$ days. The curse of dimensionality strikes early.
  \vfill
  \textbf{Grid search is impractical above 3 hyperparameters with non-trivial ranges.} The next section introduces random search, which solves this.
\end{frame}

\begin{frame}[fragile]{Code: \texttt{GridSearchCV} with Pipeline}
\begin{lstlisting}
from sklearn.model_selection import GridSearchCV
from sklearn.preprocessing import PolynomialFeatures, StandardScaler
from sklearn.linear_model import Ridge
from sklearn.pipeline import Pipeline

pipe = Pipeline([("poly", PolynomialFeatures()),
                 ("scale", StandardScaler()),
                 ("ridge", Ridge())])

grid = {
    "poly__degree": [1, 2, 3, 4, 5],
    "ridge__alpha": [1e-3, 1e-2, 1e-1, 1, 10, 100],
}

gs = GridSearchCV(pipe, grid, cv=5,
                  scoring="neg_mean_squared_error", n_jobs=-1)
gs.fit(X_train, y_train)
print(gs.best_params_, -gs.best_score_)
\end{lstlisting}
  \vfill
  \footnotesize \texttt{n\_jobs=-1} uses all CPU cores. Parameters are accessed as \texttt{step\_name\_\_param\_name}.
\end{frame}

\begin{frame}{When grid is OK}
  \begin{itemize}
    \item $\leq 2{-}3$ hyperparameters AND each has $\leq 6$ candidate values.
    \item Each model fit is cheap (small data, simple model).
    \item Reproducibility / publication: ``we evaluated all $X$ configurations.''
    \item \textbf{Discrete, narrow, well-known values} (e.g., \texttt{max\_depth} $\in \{3, 5, 7\}$ for trees) - there's nothing to randomize over.
    \item Inputs have clearly preferred discrete values from prior knowledge.
  \end{itemize}
  \vfill
  Once you have 4+ hyperparameters or 10+ values each, grid stops being practical. Random search is almost always better.
\end{frame}

% ================================================================
\section{Random search}

\begin{frame}{Why grid wastes compute in high dimensions}
  Most hyperparameters don't matter equally. For a typical Ridge + polynomial model:
  \begin{itemize}
    \item \texttt{alpha} matters a lot.
    \item \texttt{poly\_degree} matters a lot.
    \item Other small dials (e.g., \texttt{tol}, \texttt{max\_iter}) barely affect the score.
  \end{itemize}
  \vfill
  Grid search spends \emph{equal} compute on each axis. If $9$ out of $10$ axes are flat, you've used 90\% of your budget exploring flat directions.
  \vfill
  \textbf{Random search} samples each hyperparameter from a distribution. Every trial visits a new value on \emph{every} axis - including the ones that matter.
\end{frame}

\begin{frame}{Bergstra--Bengio: random $>$ grid in high dimensions}
  Classic visualization. 9 trials in 2D. Imagine the \emph{horizontal} axis matters, the \emph{vertical} axis doesn't.
  \vfill
  \begin{columns}[T]
    \column{0.5\textwidth}
    \begin{center}
      \textbf{Grid search (3x3)}
      \begin{tikzpicture}[scale=1]
        \draw[gray!50] (0,0) rectangle (3,3);
        \foreach \x in {0.5, 1.5, 2.5}
          \foreach \y in {0.5, 1.5, 2.5}
            \fill[armblue] (\x, \y) circle (3pt);
        % projection lines
        \foreach \x in {0.5, 1.5, 2.5}
          \draw[armred, dashed, thick] (\x, 0) -- (\x, -0.3);
        \node[font=\scriptsize] at (1.5, -0.7) {only 3 unique x values};
        \node[font=\tiny, anchor=east] at (0, 1.5) {y};
        \node[font=\tiny, anchor=north] at (1.5, 0) {x};
      \end{tikzpicture}
    \end{center}
    \column{0.5\textwidth}
    \begin{center}
      \textbf{Random search (9 samples)}
      \begin{tikzpicture}[scale=1]
        \draw[gray!50] (0,0) rectangle (3,3);
        \foreach \x/\y in {0.3/2.1, 0.6/0.8, 1.0/2.6, 1.3/1.4, 1.7/0.4,
                           2.0/2.2, 2.3/1.1, 2.6/2.5, 2.9/0.7}
          \fill[armblue] (\x, \y) circle (3pt);
        \foreach \x in {0.3, 0.6, 1.0, 1.3, 1.7, 2.0, 2.3, 2.6, 2.9}
          \draw[armred, dashed, thick] (\x, 0) -- (\x, -0.3);
        \node[font=\scriptsize] at (1.5, -0.7) {9 unique x values};
        \node[font=\tiny, anchor=east] at (0, 1.5) {y};
        \node[font=\tiny, anchor=north] at (1.5, 0) {x};
      \end{tikzpicture}
    \end{center}
  \end{columns}
  \vfill
  \footnotesize Random search gets 9 distinct samples along the important axis; grid gets 3. The wider the gap between effective dimensionality and total dimensionality, the bigger the random-search win.
  \vfill
  \fcolorbox{armred}{armred!8}{\parbox{0.96\textwidth}{\small
    \textbf{Default: random search for $\geq 3$ hyperparameters.} With $\sim 60$ trials, random search typically matches a $10^d$ grid. Reach for grid only when you have specific known values.
  }}
\end{frame}

\begin{frame}{Random search advantage in one line}
  If only $d_{\text{eff}}$ of your $d$ hyperparameters matter (and you don't know which), random search with $n$ trials finds a near-optimal value on each important axis - because every trial is a fresh sample on every axis.
  \vfill
  \textbf{Rule of thumb:} for $d \geq 4$ hyperparameters, random search with $60$ trials typically matches or beats grid search with $10^d$ trials.
  \vfill
  \textbf{Bonus benefits:}
  \begin{itemize}
    \item You can stop anytime (early-stopping the search). Grid forces you to finish.
    \item Easy to bound budget: set \texttt{n\_iter}.
    \item Mixes naturally with continuous distributions (\texttt{loguniform}, \texttt{uniform}, \texttt{normal}).
  \end{itemize}
\end{frame}

\begin{frame}[fragile]{Code: \texttt{RandomizedSearchCV}}
\begin{lstlisting}
from sklearn.model_selection import RandomizedSearchCV
from scipy.stats import loguniform, randint

dist = {
    "ridge__alpha":   loguniform(1e-4, 1e3),   # continuous, log-scale
    "poly__degree":   randint(1, 7),           # discrete 1..6
}

rs = RandomizedSearchCV(pipe, dist,
        n_iter=60,                  # budget
        cv=5,
        scoring="neg_mean_squared_error",
        random_state=509,           # reproducibility
        n_jobs=-1)
rs.fit(X_train, y_train)
print(rs.best_params_, -rs.best_score_)
\end{lstlisting}
  \vfill
  \scriptsize Use \texttt{loguniform} for orders-of-magnitude HPs; \texttt{uniform} for bounded ranges; \texttt{randint} for discrete. Modern alternative: \texttt{HalvingRandomSearchCV} (sklearn-builtin Successive Halving).
\end{frame}

% ================================================================
\section{Bayesian optimization}

\begin{frame}{Bayesian optimization: learn from each evaluation}
  Random search treats each trial as independent. But you've \emph{seen} the previous scores - you could pick the next trial based on what looks promising.
  \vfill
  Bayesian optimization (BO):
  \begin{enumerate}
    \item Fit a probabilistic \textbf{surrogate model} to the (hyperparameter $\to$ score) pairs seen so far. Typically a Gaussian Process or Tree-structured Parzen Estimator (TPE).
    \item Use the surrogate's mean and uncertainty to compute an \textbf{acquisition function} that scores ``how good would it be to evaluate here?''
    \item Pick the next hyperparameter as the maximum of the acquisition function.
    \item Evaluate, add to history, update surrogate. Repeat.
  \end{enumerate}
  \vfill
  Each trial uses the information from all previous trials. Random search doesn't.
\end{frame}

\begin{frame}{The surrogate + acquisition picture}
  \begin{center}
    \begin{tikzpicture}
      \begin{axis}[
        width=0.85\textwidth, height=4.5cm,
        xlabel={hyperparameter $\lambda$}, ylabel={validation score},
        xlabel style={font=\footnotesize}, ylabel style={font=\footnotesize},
        xtick=\empty, ytick=\empty,
        axis line style={gray!60}, grid=major, grid style={dashed,gray!30},
        xmin=0, xmax=6, ymin=-1.5, ymax=5.0,
        legend style={font=\scriptsize, at={(0.02,0.02)}, anchor=south west},
      ]
        % posterior mean (smooth)
        \addplot[paramgreen, very thick, smooth, name path=mean] coordinates {
          (0,-1.0) (0.5,0.0) (1.0,1.3) (1.5,1.8) (2.0,2.7) (2.5,3.5)
          (3.0,3.9) (3.5,3.5) (4.0,2.5) (4.5,1.8) (5.0,0.7) (5.5,0.0) (6.0,-0.8)};
        \addlegendentry{surrogate (posterior mean)}
        % upper / lower bounds for confidence band
        \path[name path=upper] (0,0.2) -- (0.5,0.4) -- (1.0,1.8) -- (1.5,2.0) -- (2.0,3.6) -- (2.5,3.9) -- (3.0,4.7) -- (3.5,3.9) -- (4.0,3.4) -- (4.5,2.0) -- (5.0,1.7) -- (5.5,0.4) -- (6.0,0.5);
        \path[name path=lower] (0,-2.2) -- (0.5,-0.4) -- (1.0,0.8) -- (1.5,1.6) -- (2.0,1.8) -- (2.5,3.1) -- (3.0,3.1) -- (3.5,3.1) -- (4.0,1.6) -- (4.5,1.6) -- (5.0,-0.3) -- (5.5,-0.4) -- (6.0,-2.1);
        \addplot[fill=armblue!20, draw=none] fill between[of=upper and lower];
        \addlegendentry{95\% uncertainty}
        % evaluated points
        \addplot[only marks, mark=*, mark size=2.5pt, armred] coordinates {(0.5,0.0) (1.5,1.8) (2.5,3.5) (3.5,3.5) (4.5,1.8) (5.5,0.0)};
        \addlegendentry{evaluated trials}
        % next eval marker
        \draw[->, armorange, very thick] (axis cs:3.0, 4.8) -- (axis cs:3.0, 4.0);
        \node[armorange, font=\tiny, anchor=south] at (axis cs:3.0, 4.8) {next: max acquisition};
      \end{axis}
    \end{tikzpicture}
  \end{center}
  \vfill
  \footnotesize The surrogate is confident near observed points (narrow band) and uncertain elsewhere (wide band). The acquisition function trades off \emph{exploit} (go where the mean is high) and \emph{explore} (go where the uncertainty is high). The next slide gives three concrete acquisition functions.
\end{frame}

\begin{frame}{Acquisition functions}
  Three popular acquisition functions:
  \vfill
  \begin{itemize}
    \item \textbf{Expected Improvement (EI).} How much better do we expect to do here than at the best-seen point? Smooth tradeoff between mean and uncertainty. Default in most libraries.
    \item \textbf{Upper Confidence Bound (UCB).} mean $+\, \kappa \cdot$ std. The constant $\kappa$ explicitly controls explore-vs-exploit. Used in multi-armed-bandit theory.
    \item \textbf{Probability of Improvement (PI).} Probability that the score here exceeds the current best. Greedy - tends to under-explore.
  \end{itemize}
  \vfill
  The acquisition function is itself optimized over $\lamv$ (cheap - no model fits required), then we evaluate the model only at the chosen point.
  \vfill
  \fcolorbox{armred}{armred!8}{\parbox{0.96\textwidth}{\small
    \textbf{Default: EI.} UCB if you want explicit control over the explore-vs-exploit dial. PI rarely - it under-explores. Optuna's TPE uses an EI-style criterion behind the scenes.
  }}
\end{frame}

\begin{frame}[fragile]{Code: Optuna (modern default)}
\begin{lstlisting}
import optuna
from sklearn.model_selection import cross_val_score

def objective(trial):
    alpha = trial.suggest_float("alpha", 1e-4, 1e3, log=True)
    degree = trial.suggest_int("degree", 1, 6)

    pipe = make_pipeline(PolynomialFeatures(degree),
                         StandardScaler(),
                         Ridge(alpha=alpha))
    score = cross_val_score(pipe, X_train, y_train,
                            cv=5, scoring="neg_mean_squared_error").mean()
    return score                       # Optuna maximizes by default

study = optuna.create_study(direction="maximize")
study.optimize(objective, n_trials=60)
print(study.best_params, study.best_value)
\end{lstlisting}
  \vfill
  \footnotesize Default sampler: TPE (Tree-structured Parzen Estimator). Free pruning + dashboard.
\end{frame}

\begin{frame}{When Bayesian opt is worth it}
  \textbf{Use Bayesian opt when:}
  \begin{itemize}\small
    \item Each evaluation is expensive (training takes minutes-hours, not seconds).
    \item You have $\geq 4$ hyperparameters with non-trivial interactions.
    \item You want fewer total evaluations than random search would need.
    \item You can keep evaluations sequential (Bayesian opt is inherently sequential, though batch variants exist).
  \end{itemize}
  \vfill
  \textbf{Stick with random search when:}
  \begin{itemize}\small
    \item Evaluations are cheap (seconds) - the overhead of fitting a surrogate isn't justified.
    \item You can run hundreds of parallel evaluations on a cluster - random is embarrassingly parallel; Bayesian is mostly sequential.
    \item Your search space is largely categorical with no obvious order.
  \end{itemize}
  \vfill
  \fcolorbox{armred}{armred!8}{\parbox{0.96\textwidth}{\small
    \textbf{Rule of thumb:} Bayesian opt pays off when each evaluation costs $\gtrsim 1$ minute AND you have $\geq 4$ hyperparameters. Below either threshold, the surrogate-fitting overhead isn't justified - use random search.
  }}
  \vfill
  \emph{Modern AutoML tools (Optuna, Hyperopt, Ray Tune) default to TPE-style Bayesian samplers because they ``just work'' across mixed search spaces.}
\end{frame}

% ================================================================
\section{Wrap-up}

\begin{frame}{The compute budget reality}
  Hyperparameter tuning consumes most of an ML project's compute. Practical patterns:
  \vfill
  \begin{itemize}
    \item \textbf{Subsample.} Tune on 10\% of training data, retrain best config on full data. Often within 1\% of fully-tuned.
    \item \textbf{Early stopping.} For boosted trees and NNs, stop training early when validation stops improving. Cuts each trial's cost by 2-5x.
    \item \textbf{Pruning.} Stop trials that look bad after a few epochs (Optuna does this for you).
    \item \textbf{Multi-fidelity methods.} Hyperband / Successive Halving: run many trials cheaply, gradually allocate more budget to the survivors.
    \item \textbf{Stop when score plateaus.} Track best-so-far; if it hasn't improved in 20 trials, you're done.
  \end{itemize}
  \vfill
  \fcolorbox{armred}{armred!8}{\parbox{0.96\textwidth}{\small
    \textbf{Honest reporting:} the score you tuned to optimize is \emph{optimistically biased}. Report the test-set score after tuning is complete (or use nested CV - L05b). \emph{Conceptually: HP tuning navigates the bias-variance tradeoff (L01d2). Each HP setting moves the model along the U-shape; CV finds the sweet spot empirically.}
  }}
\end{frame}

\begin{frame}{Decision flowchart}
  \begin{center}
    \begin{tikzpicture}[
      node distance=0.45cm and 0.7cm,
      box/.style={rectangle, draw=armblue, very thick, text width=2.9cm, align=center, font=\scriptsize, rounded corners=2pt, fill=armblue!5, minimum height=0.85cm},
      result/.style={rectangle, draw=paramgreen, very thick, text width=2.7cm, align=center, font=\scriptsize, rounded corners=2pt, fill=paramgreen!10, minimum height=0.85cm},
      arr/.style={->, >=stealth, thick, armblue},
    ]
      \node[box] (start) {How many hyperparameters?};
      \node[result, right=of start] (manual) {Manual or grid (1-2 HPs)};
      \node[box, below=of start] (expensive) {Each evaluation expensive (mins+)?};
      \node[result, right=of expensive] (bo) {Bayesian opt (Optuna TPE)};
      \node[box, below=of expensive] (parallel) {Heavily parallel hardware?};
      \node[result, right=of parallel] (rs) {Random search};
      \node[result, below=of parallel] (rs2) {Random search (default)};

      \draw[arr] (start) -- node[above, font=\tiny]{$\leq 2$} (manual);
      \draw[arr] (start) -- node[left, font=\tiny]{$\geq 3$} (expensive);
      \draw[arr] (expensive) -- node[above, font=\tiny]{yes} (bo);
      \draw[arr] (expensive) -- node[left, font=\tiny]{no} (parallel);
      \draw[arr] (parallel) -- node[above, font=\tiny]{yes} (rs);
      \draw[arr] (parallel) -- node[left, font=\tiny]{no} (rs2);
    \end{tikzpicture}
  \end{center}
  \vfill
  \footnotesize Defaults: $\leq 2$ HPs $\to$ grid is fine; otherwise random search; if compute is the bottleneck, Bayesian opt. \\[2pt]
  \textbf{Whatever method:} use subsampling, early stopping, and pruning to multiply your effective budget (see compute-budget frame).
\end{frame}

\begin{frame}{Recap}
  \begin{itemize}\small
    \item \textbf{Parameters} are learned from data. \textbf{Hyperparameters} are chosen by you. Tuning $=$ outer optimization wrapping the inner fit.
    \item \textbf{Manual.} Cheap, intuition-building, doesn't scale. Watch for confirmation bias and validation-set abuse.
    \item \textbf{Grid.} Exhaustive over Cartesian product. Fine for $\leq 2$ HPs; explodes in high dimensions.
    \item \textbf{Random.} Default for $\geq 3$ HPs. Each trial is a fresh sample on every axis. Easy to budget and parallelize. Beats grid when effective dimensionality is low.
    \item \textbf{Bayesian.} Models the score-vs-HP surface and queries smartly. Best when each evaluation is expensive. Optuna's TPE is the modern default; \texttt{HalvingRandomSearchCV} is the sklearn-builtin alternative.
    \item \textbf{Bias-variance lens (L01d2):} HP tuning navigates the bias-variance U. Each candidate setting moves the model along the U; CV finds the sweet spot. Without this lens HP tuning feels arbitrary; with it, it's purposeful.
    \item \textbf{Honest reporting.} Tuning inflates the score; report on a held-out test set or use nested CV (L05b).
  \end{itemize}
\end{frame}

\begin{frame}{HW01e - practice}
  \footnotesize Estimated time: 90-120 minutes. Tasks 1-3 graded; bonuses optional.
  \begin{enumerate}\small
    \item On the Yerevan rent dataset: manually tune Ridge \texttt{alpha} on $\{0.001, 0.01, 0.1, 1, 10, 100, 1000\}$ using 5-fold CV. Plot validation MSE vs.\ \texttt{alpha} on a log-x axis. Report the best.
    \item Use \texttt{GridSearchCV} to tune \texttt{poly\_degree} $\in \{1,2,3,4,5\}$ and \texttt{alpha} (same grid). Report best params and wall-clock time.
    \item Replace step 2 with \texttt{RandomizedSearchCV}: \texttt{poly\_degree} $\sim$ \texttt{randint(1,6)}, \texttt{alpha} $\sim$ \texttt{loguniform(1e-4, 1e3)}, \texttt{n\_iter=60}. Compare best score and wall-clock time to grid.
    \item \emph{Bonus 1:} run Optuna with TPE for 60 trials on the same search space. Compare to random search. Plot Optuna's history.
    \item \emph{Bonus 2:} extend to XGBoost with 5 hyperparameters. Run grid (cap at 10 minutes), random ($n\_iter=60$), and Optuna ($n\_trials=60$). Which wins on a fixed time budget? \textbf{Most instructive question.}
    \item \emph{Bonus 3:} wrap your best pipeline in nested CV (L05b) and report an honest generalization score. How much did tuning inflate the optimistic estimate?
  \end{enumerate}
\end{frame}

\end{document}

% =============================================================
% Provenance:
%   - Topics: manual, grid, random, Bayesian HP tuning.
%   - Follows L01d evaluation protocol naturally: CV is the scorer
%     that all tuning methods rely on.
%   - Yerevan running example continues.
%   - Forward reference to L05b nested CV.
%   - Math module 14 (Bayesian optimization) cross-reference implicit.
%   - Predict-first frames per memory feedback_predict_first_slides.md.
% =============================================================
