\documentclass[aspectratio=169]{beamer}
\input{../preamble}

\usepgfplotslibrary{groupplots, fillbetween}
\usepackage{listings}

\lstset{
  basicstyle=\ttfamily\footnotesize,
  keywordstyle=\color{armblue}\bfseries,
  commentstyle=\color{paramgreen}\itshape,
  stringstyle=\color{armred},
  showstringspaces=false,
  language=Python,
  frame=single,
  framerule=0pt,
  backgroundcolor=\color{lightbg},
  numbers=none,
  breaklines=true,
  xleftmargin=4pt,
  xrightmargin=4pt,
}

\title{L01g: Feature Engineering}
\subtitle{Where most of an ML project's time actually goes}
\date{}

\begin{document}

\begin{frame}{Outline}
  \tableofcontents
\end{frame}

% ================================================================
\section{What and why}

\begin{frame}[fragile]{What is feature engineering?}
  Three related-but-distinct activities on tabular data:
  \vfill
  \begin{itemize}\footnotesize
    \item \textbf{Preprocessing (L01c)} - \emph{Fix} the data. Same columns, cleaner. \\
{\lstset{basicstyle=\ttfamily\scriptsize}\begin{lstlisting}
df["rent"] = df["rent"].fillna(df["rent"].median())
\end{lstlisting}}
    \item \textbf{Feature engineering} - \emph{Create} new columns. More columns. \\
{\lstset{basicstyle=\ttfamily\scriptsize}\begin{lstlisting}
df["price_per_m2"] = df["rent"] / df["area"]
\end{lstlisting}}
    \item \textbf{Feature selection (L01h)} - \emph{Choose} which to keep. Fewer columns. \\
{\lstset{basicstyle=\ttfamily\scriptsize}\begin{lstlisting}
X = X[["price_per_m2", "rooms", "district_te"]]
\end{lstlisting}}
  \end{itemize}
  \vfill
  \footnotesize \textbf{The classical-ML maxim:} ``Good features beat a better algorithm.'' Most Kaggle winners come from FE, not the model class. \emph{Deep learning} learns features from raw inputs (pixels, tokens, audio); for tabular data, hand-engineered features still dominate.
\end{frame}

\begin{frame}{Predict-first: where does an ML project's time go?}
  An ML project has several stages: data collection, EDA, preprocessing, feature engineering, modeling, tuning, evaluation, deployment.
  \vfill
  \pause
  \textbf{Predict: what fraction of project time is spent on data \& features vs modeling \& tuning?}
  \pause
  \vfill
  \begin{columns}[T]
    \column{0.55\textwidth}
    Surveys (CrowdFlower 2016, Anaconda 2020) consistently report:
    \begin{itemize}\small
      \item \textbf{60-80\%} on data collection, cleaning, feature engineering.
      \item \textbf{10-20\%} on modeling and tuning.
      \item \textbf{5-15\%} on deployment and monitoring.
    \end{itemize}
    \vskip 4pt
    \footnotesize Students often expect the reverse. The reality: \emph{the algorithm is the easy part}; the data preparation is the work.
    \column{0.42\textwidth}
    \begin{center}
      \begin{tikzpicture}
        \begin{axis}[width=\textwidth, height=4.2cm,
                     xbar, xmin=0, xmax=80,
                     symbolic y coords={deploy, tuning, modeling, data+features},
                     ytick=data,
                     xlabel={\% of project time}, xlabel style={font=\tiny},
                     yticklabel style={font=\tiny},
                     xticklabel style={font=\tiny},
                     nodes near coords, nodes near coords style={font=\tiny},
                     enlarge y limits=0.2,
                     axis line style={gray!60}]
          \addplot[fill=armblue!50, draw=armblue] coordinates {
            (65,data+features) (15,modeling) (10,tuning) (10,deploy)};
        \end{axis}
      \end{tikzpicture}
    \end{center}
  \end{columns}
\end{frame}

\begin{frame}{Sources of new features}
  Three super-categories. Yerevan rent examples in italics.
  \vfill
  \begin{columns}[T]
    \column{0.33\textwidth}
    \footnotesize
    \textbf{From the same row}\\
    (one observation in, one feature out)
    \begin{itemize}\scriptsize
      \item Transformations: log, power, binning. \emph{log(rent)}
      \item Combinations: ratios, differences, products. \emph{rent / area, year\_listed $-$ year\_built}
      \item Polynomial / interactions. \emph{area $\times$ floor}
    \end{itemize}
    \column{0.33\textwidth}
    \footnotesize
    \textbf{From many rows}\\
    (group of observations $\to$ feature)
    \begin{itemize}\scriptsize
      \item Group-by aggregations. \emph{avg\_rent\_by\_district}
      \item Rolling windows (trailing). \emph{mean rent over last 30 days}
      \item Lags. \emph{rent at the previous listing}
      \item Density / k-NN. \emph{listings within 1 km}
    \end{itemize}
    \column{0.33\textwidth}
    \footnotesize
    \textbf{From outside the data}\\
    (joins to external sources)
    \begin{itemize}\scriptsize
      \item Geospatial joins. \emph{distance to Republic Square}
      \item Temporal joins. \emph{is\_holiday(listed\_at)}
      \item Demographic joins. \emph{district median income}
      \item Domain reference data. \emph{m\textsuperscript{2} prices from official index}
    \end{itemize}
  \end{columns}
  \vfill
  \footnotesize \textbf{Domain knowledge cuts across all three} - the biggest single multiplier. The rest of this deck walks through each category.
\end{frame}

% ================================================================
\section{Numerical and transformation features}

\begin{frame}[fragile]{Numerical transformations}
  \begin{columns}[T]
    \column{0.55\textwidth}
    \footnotesize
    \textbf{Log / power.} For skewed positive targets (recap L01c). \texttt{np.log1p(price)} compresses the tail.
    \vskip 4pt
    \textbf{Binning.} Convert a continuous feature into categories. Three strategies:
    \begin{itemize}\footnotesize
      \item Equal-width: same range per bin.
      \item Equal-frequency (quantile): same count per bin.
      \item Supervised: pick splits that maximize separation in $y$.
    \end{itemize}
    \vskip 4pt
    \textbf{When to bin?} When the target depends \emph{non-monotonically} on the feature (e.g., rent peaks at 60 m$^2$ then plateaus) and you're using a linear model. Trees don't need binning.
    \column{0.45\textwidth}
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.preprocessing \
  import KBinsDiscretizer

# 5 quantile bins on area
binner = KBinsDiscretizer(
    n_bins=5,
    encode="ordinal",
    strategy="quantile")
df["area_bin"] = binner.fit_transform(
    df[["area"]])
\end{lstlisting}}
    \vskip 4pt
    \begin{center}
      \begin{tikzpicture}[scale=0.7]
        \draw[gray!50, thick] (0,0) -- (8,0);
        \foreach \x in {0,2,4,6,8}
          \draw[gray!50] (\x, 0.1) -- (\x, -0.1);
        \foreach \i/\lbl in {1/B1, 3/B2, 5/B3, 7/B4} {
          \node[font=\tiny, armblue] at (\i, -0.4) {\lbl};
        }
        \foreach \x in {0.6, 1.2, 1.8, 2.4, 3.0, 3.6, 4.2, 4.8, 5.4, 6.0, 6.6, 7.2}
          \fill[armred] (\x, 0.3) circle (1pt);
        \node[font=\scriptsize, anchor=south] at (4, 0.6) {continuous $\to$ 4 bins};
      \end{tikzpicture}
    \end{center}
  \end{columns}
\end{frame}

\begin{frame}{Polynomial and interaction features}
  Linear regression on \emph{engineered} features is the most underrated trick in classical ML. From L01b: polynomial regression is OLS with $\phi(x) = (1, x, x^2, \ldots, x^d)$.
  \vfill
  \textbf{Interactions} extend this to multiple features. Given $x_1, x_2$:
  \[
    \phi(\xv) = (1,\; x_1,\; x_2,\; x_1^2,\; x_2^2,\; x_1 x_2)
  \]
  The product $x_1 x_2$ lets a linear model express ``effect of $x_1$ depends on level of $x_2$.''
  \vfill
  \fcolorbox{armred}{armred!8}{\parbox{0.96\textwidth}{\small
    \textbf{Predict-first:} you expect rent to depend on \texttt{area} $\times$ \texttt{floor}. Do you need to add the product explicitly for: (a) Ridge regression, (b) Random Forest?
    \pause
    \\[2pt]
    Answers: (a) \textbf{Yes} - a linear model can't construct products. (b) \textbf{No} - trees discover interactions via nested splits. Polynomial features are a \emph{linear-model crutch}, not a universal good.
  }}
\end{frame}

\begin{frame}[fragile]{Ratios and differences}
  Many of the most predictive features are simple ratios or differences. Linear models can't construct them; trees can but slowly.
  \vfill
  \begin{columns}[T]
    \column{0.5\textwidth}
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
# Yerevan-rent examples
df["price_per_m2"]    = df["rent"] / df["area"]
df["bedroom_density"] = df["rooms"] / df["area"]
df["age_at_listing"]  = (df["year_listed"]
                         - df["year_built"])

# Classical examples
df["bmi"]   = df["weight"] / df["height"]**2
df["lti"]   = df["loan"] / df["income"]
df["debt_to_equity"] = df["debt"] / df["equity"]
\end{lstlisting}}
    \column{0.5\textwidth}
    \footnotesize
    \textbf{Why ratios beat raw features:}
    \begin{itemize}\footnotesize
      \item Often more predictive than either component alone.
      \item Express the underlying \emph{rate} or \emph{intensity}.
      \item Scale-invariant - a 3000 USD loan to a 6000 USD earner is the same as 30k/60k.
      \item Sparse - one engineered feature vs.\ two raw ones.
    \end{itemize}
    \textbf{When NOT to engineer:}
    \begin{itemize}\footnotesize
      \item If the denominator can be 0 or negative.
      \item For deep learning on tabular - just feed in the raw inputs.
    \end{itemize}
  \end{columns}
\end{frame}

\begin{frame}[fragile]{Datetime features}
  \begin{columns}[T]
    \column{0.6\textwidth}
    \footnotesize Datetimes are rich. Extract components, time-since-events, calendar markers, and cyclic encodings.
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
df["dt"] = pd.to_datetime(df["listed_at"])
df["year"]    = df["dt"].dt.year
df["month"]   = df["dt"].dt.month
df["weekday"] = df["dt"].dt.dayofweek    # 0=Mon
df["is_weekend"]    = df["weekday"].isin([5, 6])
df["days_since_listed"] = (today - df["dt"]).dt.days
df["is_holiday"] = df["dt"].dt.date.isin(holidays)

# cyclic encoding for periodic features
df["hour_sin"] = np.sin(2*np.pi * df["hour"]/24)
df["hour_cos"] = np.cos(2*np.pi * df["hour"]/24)
\end{lstlisting}}
    \footnotesize Watch \textbf{leakage} on ``time since last event'' - use events strictly before the row's timestamp.
    \column{0.4\textwidth}
    \begin{center}
      \textbf{\footnotesize Why cyclic encoding?}\\[2pt]
      \scriptsize Hour 23 and hour 0 are 1 hour apart, not 23. Map onto the unit circle:\\[4pt]
      \begin{tikzpicture}[scale=1.5]
        \draw[gray!50, thick] (0,0) circle (1);
        \draw[gray!40, ->] (-1.2,0) -- (1.2,0) node[right, font=\tiny] {cos};
        \draw[gray!40, ->] (0,-1.2) -- (0,1.2) node[above, font=\tiny] {sin};
        % hours
        \foreach \h/\a in {0/90, 6/0, 12/-90, 18/180} {
          \fill[armblue] ({cos(\a)}, {sin(\a)}) circle (1.2pt);
          \node[font=\tiny, armblue] at ({1.25*cos(\a)}, {1.25*sin(\a)}) {\h};
        }
        % hour 23 (close to 0)
        \fill[armred] ({cos(90 + 15)}, {sin(90 + 15)}) circle (1.5pt);
        \node[font=\tiny, armred] at ({1.3*cos(90+15)}, {1.3*sin(90+15)}) {23};
        \draw[armorange, very thick, <->] ({cos(90)}, {sin(90)}) to[bend right=20] ({cos(105)}, {sin(105)});
        \node[font=\tiny, armorange, anchor=south] at (0.1, 1.15) {close!};
      \end{tikzpicture}\\[2pt]
      \scriptsize Hour 23 (red) lands right next to hour 0 (blue) on the circle.
    \end{center}
  \end{columns}
\end{frame}

% ================================================================
\section{Aggregations and time-series}

\begin{frame}[fragile]{Group-by aggregations}
  For each row, attach summary statistics of its group. Hugely effective for tabular ML.
  \vfill
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
# Yerevan: for each apartment, learn the district's price profile
df["avg_rent_district"]   = df.groupby("district")["rent"].transform("mean")
df["med_rent_district"]   = df.groupby("district")["rent"].transform("median")
df["std_rent_district"]   = df.groupby("district")["rent"].transform("std")
df["count_in_district"]   = df.groupby("district")["rent"].transform("count")
df["rank_in_district"]    = df.groupby("district")["rent"].rank(pct=True)
df["rent_vs_district_avg"] = df["rent"] / df["avg_rent_district"]
\end{lstlisting}}
  \vfill
  \fcolorbox{armred}{armred!8}{\parbox{0.96\textwidth}{\small
    \textbf{Leakage warning:} computing group means on the \emph{full} dataset before splitting leaks the target into features. \emph{Typical inflation in our Yerevan example:} honest CV MAE $= 32$ kAMD; leaky CV MAE $= 24$ kAMD - the model looks 25\% better than it actually is. Fit aggregations on the train fold only, or use a CV-aware encoder from \texttt{category\_encoders}.
  }}
\end{frame}

\begin{frame}[fragile]{Time-series features: lags and rolling windows}
  \begin{columns}[T]
    \column{0.55\textwidth}
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
df = df.sort_values("dt")

# lag features
df["lag_1"]  = df["price"].shift(1)
df["lag_7"]  = df["price"].shift(7)

# rolling windows (trailing!)
df["ma_7"]  = df["price"].rolling(7).mean()
df["std_7"] = df["price"].rolling(7).std()

# expanding / EMA
df["ema_7"] = df["price"].ewm(span=7).mean()

# diff features
df["diff_1"]   = df["price"].diff(1)
df["pct_chg"]  = df["price"].pct_change()
\end{lstlisting}}
    \column{0.45\textwidth}
    \begin{center}
      \begin{tikzpicture}
        \begin{axis}[width=\textwidth, height=3.5cm,
                     xlabel={time $t$}, ylabel={price},
                     xlabel style={font=\tiny}, ylabel style={font=\tiny},
                     xtick=\empty, ytick=\empty, axis line style={gray!60}]
          \addplot[armblue, very thick, mark=*, mark size=1pt]
            coordinates {(1,3) (2,3.2) (3,3.5) (4,3.8) (5,3.6) (6,4.2)
                         (7,4.5) (8,4.7) (9,4.4) (10,5.0)};
          \draw[armred, very thick, |-|] (axis cs:4,1.5) -- (axis cs:10,1.5);
          \node[font=\tiny, armred] at (axis cs:7, 1.2) {trailing window};
          \draw[armorange, very thick, ->] (axis cs:10.5, 5.0) -- (axis cs:11, 5.0);
          \node[font=\tiny, armorange] at (axis cs:11, 4.5) {predict};
        \end{axis}
      \end{tikzpicture}
    \end{center}
    \vskip 4pt
    \fcolorbox{armred}{armred!8}{\parbox{0.96\textwidth}{\scriptsize
      \textbf{Look-ahead bias:} \texttt{rolling(7, center=True)} silently uses future values. \emph{Never} center rolling windows when forecasting.
    }}
  \end{columns}
\end{frame}

\begin{frame}[fragile]{Geospatial features}
  When you have lat/lon, the raw coordinates are usually useless; the \emph{relationships} are predictive.
  \vfill
  \begin{columns}[T]
    \column{0.55\textwidth}
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from math import radians, sin, cos, asin, sqrt

def haversine(lat1, lon1, lat2, lon2):
    R = 6371      # km
    dlat = radians(lat2 - lat1)
    dlon = radians(lon2 - lon1)
    a = (sin(dlat/2)**2
         + cos(radians(lat1))
         * cos(radians(lat2))
         * sin(dlon/2)**2)
    return 2 * R * asin(sqrt(a))

# Distance to Republic Square (Yerevan center)
df["dist_to_center"] = df.apply(lambda r:
    haversine(r.lat, r.lon, 40.177, 44.512),
    axis=1)
\end{lstlisting}}
    \column{0.45\textwidth}
    \footnotesize
    Useful geospatial features:
    \begin{itemize}\footnotesize
      \item \textbf{Distances} to landmarks (metro, river, parks).
      \item \textbf{Cluster IDs} from k-means or DBSCAN on (lat, lon).
      \item \textbf{Grid cells} via geohash, H3, or S2 - aggregations within a cell.
      \item \textbf{Density features} (k-NN density estimate around each point).
      \item \textbf{POI counts} within a radius (cafes, schools, public transport).
    \end{itemize}
    \vskip 4pt
    Raw lat/lon as columns in a linear model: don't. Trees can split on lat/lon but distances/clusters are usually more efficient.
  \end{columns}
\end{frame}

\begin{frame}[fragile]{Text features (intro)}
  Text needs to become numbers. Three layers, increasing in power:
  \vfill
  \begin{itemize}
    \item \textbf{Engineered scalars.} Length, word count, punctuation count, ALL-CAPS ratio, emoji count. Cheap and surprisingly predictive for spam, sentiment.
    \item \textbf{Bag-of-words / TF-IDF / n-grams.} Tokenize, count, weight by inverse document frequency. The classical NLP workhorse.
    \item \textbf{Embeddings.} Word2Vec, sentence-BERT, modern transformer encoders. Deferred to the NLP chapter.
  \end{itemize}
  \vfill
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.feature_extraction.text import TfidfVectorizer
tfidf = TfidfVectorizer(max_features=2000, ngram_range=(1, 2),
                        min_df=5, stop_words="english")
X_text = tfidf.fit_transform(df["description"])

df["desc_length"]  = df["description"].str.len()
df["caps_ratio"]   = df["description"].str.count(r"[A-Z]") / df["desc_length"]
df["num_exclaim"]  = df["description"].str.count(r"!")
\end{lstlisting}}
\end{frame}

\begin{frame}[fragile]{Feature crosses for high-cardinality}
  When two high-cardinality categoricals matter \emph{jointly}, you can either combine them into one feature (a ``cross'') or learn an embedding.
  \vfill
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
# Manual cross: district x rooms -> Kentron_3br
df["district_x_rooms"] = df["district"] + "_" + df["rooms"].astype(str)

# Then encode the cross with target / frequency / OHE
\end{lstlisting}}
  \vfill
  \begin{itemize}
    \item Used heavily in classical CTR prediction (Google's Wide \& Deep paper, 2016).
    \item Risk: explosion. 100 districts $\times$ 10 room counts $=$ 1000 categories.
    \item Modern alternative: learn an \emph{embedding} per categorical, sum or concatenate. NN does this naturally; \texttt{category\_encoders} has \texttt{HashingEncoder} for the classical version.
  \end{itemize}
\end{frame}

% ================================================================
\section{Selection preview}

\begin{frame}[fragile]{After engineering: select (L01h preview)}
  More features is not always better. \emph{Curse of dimensionality} (math 30), multicollinearity (L01b), variance growth, compute cost, and interpretability all push for fewer features.
  \vfill
  \textbf{Three families} (full treatment in L01h):
  \begin{itemize}\footnotesize
    \item \textbf{Filter} - score each feature independently (\texttt{f\_regression}, mutual info). Fast but misses interactions.
    \item \textbf{Embedded} - the model selects while fitting. \emph{Lasso}, tree importances, permutation importance.
    \item \textbf{Wrapper} - search subsets via CV. \emph{RFE-CV}, sequential forward/backward, Boruta. Slow but accurate.
  \end{itemize}
  \vfill
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.feature_selection import SelectKBest, f_regression
from sklearn.linear_model import LassoCV

# Filter: top 20 by linear correlation with y
SelectKBest(f_regression, k=20).fit(X, y)
# Embedded: L1 forces sparsity
lasso = LassoCV(cv=5).fit(X, y)        # zero coefficients = dropped
\end{lstlisting}}
  \vfill
  \footnotesize \textbf{Workflow:} filter for initial cut ($p > 10{,}000 \to 500$), embedded for production ($500 \to 50$), wrapper only when stakes justify the compute. L01h covers each family in depth.
\end{frame}

% ================================================================
\section{Automation, pitfalls, production}

\begin{frame}[fragile]{Automated feature engineering}
  Tools that generate features automatically from your raw data:
  \vfill
  \begin{itemize}
    \item \textbf{featuretools.} ``Deep Feature Synthesis'' walks relational schemas and generates aggregations, transformations, and combinations.
    \item \textbf{tsfresh.} Extracts hundreds of statistical features from time-series.
    \item \textbf{autofeat, Featurewiz.} General-purpose automated FE.
  \end{itemize}
  \vfill
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
import featuretools as ft
es = ft.EntitySet(id="rent")
es = es.add_dataframe(dataframe=df, dataframe_name="listings", index="id")
es = es.add_dataframe(dataframe=districts, dataframe_name="districts", ...)
feature_matrix, defs = ft.dfs(entityset=es,
                              target_dataframe_name="listings",
                              max_depth=2)
\end{lstlisting}}
  \vfill
  \footnotesize \textbf{What DFS produces (concrete).} Given table \texttt{listings} with foreign key to \texttt{landlords}, DFS generates: \texttt{MEAN(landlords.rating)}, \texttt{COUNT(landlords.listings)}, \texttt{STD(landlords.rent)}, \texttt{MAX(landlords.years\_active)}, plus transformations like \texttt{DAY(dt)}, \texttt{MONTH(dt)}. Hundreds of candidates from a small schema.\\[2pt]
  \textbf{When to use:} early exploration, complex relational data, generating candidates.\\
  \textbf{When NOT to use:} production (hard to maintain, slow at inference). Auto-generated features usually \emph{lose} to a handful of domain features.
\end{frame}

\begin{frame}{Six feature-engineering pitfalls}
  \begin{enumerate}\small
    \item \textbf{Target leakage.} Computing a target-derived feature (mean rent by district) on the \emph{full} dataset before splitting. Use train fold only or a CV-aware encoder.
    \item \textbf{Look-ahead bias.} For time-series, using future values in a ``rolling'' or aggregate feature. Always trailing windows; never \texttt{center=True}.
    \item \textbf{Train-serve skew.} A feature easy to compute on training data but impossible at inference (e.g., ``count of total purchases'' when the user's full history isn't known yet at inference time).
    \item \textbf{Hardcoded statistics.} ``Just divide by the mean rent'' - whose mean? Save the fitted statistic via a Pipeline (L01c); don't recompute on test.
    \item \textbf{Over-engineering.} 500 features on 1000 rows. Curse of dimensionality $+$ multicollinearity. Stop when CV stops improving; verify via L01h.
    \item \textbf{Engineering features the model already discovers.} Adding $x_1 x_2$ to a Random Forest is wasted compute - the tree finds that interaction via nested splits. \emph{Skip polynomial features for tree models.} Always check feature importance (L01h SHAP / permutation) before adding more.
  \end{enumerate}
\end{frame}

\begin{frame}{Feature stores: training and inference see the same features}
  In production the model you train and the model you serve see different code paths. Without care, they compute features differently.
  \vfill
  \begin{columns}[T]
    \column{0.55\textwidth}
    \footnotesize
    \textbf{A feature store:}
    \begin{itemize}\footnotesize
      \item Stores feature definitions (one function, one source of truth).
      \item Serves \textbf{batch} queries for training (historical, point-in-time).
      \item Serves \textbf{online} queries for inference (low-latency lookup).
      \item Guarantees train/serve consistency.
      \item Versions features over time.
      \item Enforces point-in-time correctness for time-series (no leakage).
    \end{itemize}
    \vskip 4pt
    Tools: \textbf{Feast} (open source), \textbf{Tecton}, \textbf{Vertex AI FS}, \textbf{SageMaker FS}.
    \column{0.45\textwidth}
    \begin{center}
      \begin{tikzpicture}[
        node distance=0.35cm and 0cm,
        defn/.style={rectangle, draw=armblue, very thick, text width=2.6cm, align=center, font=\scriptsize, rounded corners=2pt, fill=armblue!10, minimum height=0.65cm},
        store/.style={rectangle, draw=armorange, very thick, text width=2.6cm, align=center, font=\scriptsize, rounded corners=2pt, fill=armorange!15, minimum height=0.65cm},
        endpt/.style={rectangle, draw=paramgreen, very thick, text width=2.6cm, align=center, font=\scriptsize, rounded corners=2pt, fill=paramgreen!10, minimum height=0.7cm},
        arr/.style={->, >=stealth, thick, armblue},
      ]
        \node[defn] (def) {Feature definition\\(one Python function)};
        \node[store, below=of def] (st) {Feature store};
        \node[endpt, below=of st] (train) {Training\\\scriptsize batch query, historical};
        \node[endpt, below=of train] (serve) {Inference\\\scriptsize online lookup, real-time};
        \draw[arr] (def) -- (st);
        \draw[arr] (st) -- (train);
        \draw[arr] (train) -- (serve);
      \end{tikzpicture}
    \end{center}
  \end{columns}
  \vfill
  \footnotesize For coursework: a Pipeline (L01c) gets you most of the way. For deployed models with feedback loops: you need a real feature store.
\end{frame}

% ================================================================
\section{Wrap-up}

\begin{frame}{Decision flowchart}
  \begin{center}
    \begin{tikzpicture}[
      node distance=0.4cm and 0.7cm,
      box/.style={rectangle, draw=armblue, very thick, text width=3cm, align=center, font=\scriptsize, rounded corners=2pt, fill=armblue!5, minimum height=0.85cm},
      result/.style={rectangle, draw=paramgreen, very thick, text width=2.8cm, align=center, font=\scriptsize, rounded corners=2pt, fill=paramgreen!10, minimum height=0.85cm},
      arr/.style={->, >=stealth, thick, armblue},
    ]
      \node[box] (start) {Have domain knowledge?};
      \node[result, right=of start] (dom) {Start with domain features (ratios, diffs)};
      \node[box, below=of start] (linear) {Linear / kernel model?};
      \node[result, right=of linear] (poly) {Add polynomial + interactions};
      \node[box, below=of linear] (time) {Time-series target?};
      \node[result, right=of time] (lag) {Lag + rolling features};
      \node[box, below=of time] (cat) {High-cardinality categoricals?};
      \node[result, right=of cat] (te) {Target encoding (CV-aware!)};
      \node[box, below=of cat] (auto) {Many tables, exploring?};
      \node[result, right=of auto] (ft) {featuretools / tsfresh};

      \draw[arr] (start) -- node[above, font=\tiny]{yes} (dom);
      \draw[arr] (start) -- (linear);
      \draw[arr] (linear) -- node[above, font=\tiny]{yes} (poly);
      \draw[arr] (linear) -- (time);
      \draw[arr] (time) -- node[above, font=\tiny]{yes} (lag);
      \draw[arr] (time) -- (cat);
      \draw[arr] (cat) -- node[above, font=\tiny]{yes} (te);
      \draw[arr] (cat) -- (auto);
      \draw[arr] (auto) -- node[above, font=\tiny]{yes} (ft);
    \end{tikzpicture}
  \end{center}
  \vfill
  \footnotesize After engineering, always do a feature-selection pass (Lasso or tree importance) and check CV improvement before keeping.
\end{frame}

\begin{frame}{Recap}
  \begin{itemize}\small
    \item \textbf{60-80\% of an ML project} is data and features, not modeling. ``Good features beat a better algorithm'' on tabular data.
    \item \textbf{Domain knowledge is the multiplier.} Ratios (BMI, price-per-m$^2$), differences (age, days-since), and rates beat raw features.
    \item \textbf{Transformations:} log/power for skew, binning for non-monotonic effects, polynomial $+$ interactions for linear models (\emph{not for trees}).
    \item \textbf{Aggregations:} group-by stats, rolling windows (trailing only), expanding stats. The leakage hides in the fit-on-full-data step.
    \item \textbf{Specialty domains:} datetime (cyclic, components), geospatial (distances, clusters), text (TF-IDF), high-cardinality crosses.
    \item \textbf{Bias-variance lens (L01d2):} each new feature lowers bias and raises variance. Stop when CV stops improving.
    \item \textbf{Selection (L01h):} filter $\to$ embedded $\to$ wrapper, in order of cost. Always do a feature-selection pass after engineering.
    \item \textbf{Automation tools} (featuretools, tsfresh) for exploration, not production. Domain features still win.
    \item \textbf{Pitfalls:} target leakage, look-ahead bias, train-serve skew, engineering what the model already discovers. Wrap everything in a Pipeline (L01c); serve from a feature store in production.
  \end{itemize}
\end{frame}

\begin{frame}{HW01g - practice}
  \footnotesize Estimated time: 90-120 minutes. Tasks 1-4 graded; bonuses optional.
  \begin{enumerate}\footnotesize
    \item On the Yerevan rent dataset, engineer 4-5 domain features by hand (\texttt{price\_per\_m2}, \texttt{age\_at\_listing}, \texttt{avg\_rent\_district}, \texttt{rent\_vs\_district\_avg}). \emph{If lat/lon is available, add \texttt{dist\_to\_center}.} Fit Ridge with and without; report test MAE.
    \item Add \texttt{PolynomialFeatures(degree=2)} to the original feature set. Compare CV score to step 1.
    \item Fit Lasso on the union of original $+$ engineered $+$ polynomial features. Which features survive? Compare to manual-only.
    \item Replace Ridge with a tree-based model (\texttt{HistGradientBoostingRegressor}). Drop polynomial features (recall pitfall \#6). Compare to linear pipeline.
    \item \emph{Bonus 1:} compute \texttt{avg\_rent\_district} the wrong way (full data, then split). Verify the $\sim 25\%$ inflation from frame 9.
    \item \emph{Bonus 2:} use \texttt{featuretools}. Compare auto-features to manual. Which wins on test? \textbf{Most instructive.}
    \item \emph{Bonus 3:} rolling 30-day average rent feature with listings sorted by date. Verify no look-ahead bias.
    \item \emph{Bonus 4:} use SHAP / permutation importance (L01h). Drop near-zero-importance features. Does CV improve?
  \end{enumerate}
\end{frame}

\end{document}

% =============================================================
% Provenance:
%   - Topic from plan_manual.md Practical concepts item 1.
%   - Yerevan apartment running example continued from L01b-L01f.
%   - Cross-references: L01b (polynomial), L01c (preprocessing, datetime, leakage),
%     L01d (CV for unbiased reporting), math module 30 (curse of dimensionality).
%   - Predict-first frames per memory feedback_predict_first_slides.md.
% =============================================================
