\documentclass[aspectratio=169]{beamer}
\input{../preamble}

\usepgfplotslibrary{groupplots, fillbetween}
\usepackage{listings}

\lstset{
  basicstyle=\ttfamily\footnotesize,
  keywordstyle=\color{armblue}\bfseries,
  commentstyle=\color{paramgreen}\itshape,
  stringstyle=\color{armred},
  showstringspaces=false,
  language=Python,
  frame=single,
  framerule=0pt,
  backgroundcolor=\color{lightbg},
  numbers=none,
  breaklines=true,
  xleftmargin=4pt,
  xrightmargin=4pt,
}

\title{L07: Classification Metrics}
\subtitle{Accuracy is not enough - and what to use instead}
\date{}

\begin{document}

\begin{frame}{Outline}
  \tableofcontents
\end{frame}

% ================================================================
\section{Accuracy and its problem}

\begin{frame}{Setup: the cheese factory running example}
  Imagine a cheese factory producing 1000 wheels per day. About \textbf{3\% have a quality problem} - contamination, spoilage, or off-flavor. Bad cheese reaching customers causes recalls, illness, lawsuits.
  \vfill
  Binary classification task:
  \begin{itemize}
    \item \textbf{Positive class} (the rare class, $\sim 3\%$): contaminated / bad cheese batch.
    \item \textbf{Negative class} (the common class, $\sim 97\%$): clean / good cheese batch.
  \end{itemize}
  \vfill
  We build a model that scores each batch's risk -- most classifiers output a \emph{probability}, not a hard label. \emph{(That score is more useful than the label; we'll see why halfway through this deck.)} We need:
  \begin{enumerate}\small
    \item A metric to \emph{compare} candidate models.
    \item A way to pick a \emph{threshold} that turns scores into decisions.
    \item A method that respects the \emph{asymmetric cost} (missing bad cheese is much worse than over-inspecting clean cheese).
  \end{enumerate}
  \vfill
  This deck walks through what works, what doesn't, and why. The cascade: accuracy $\to$ precision/recall $\to$ ROC AUC $\to$ PR AUC, each one solving the previous one's problem.
\end{frame}

\begin{frame}[fragile]{Accuracy: the obvious starting point}
  \[
    \text{Accuracy} = \frac{\text{correct predictions}}{\text{total predictions}}
  \]
  \vfill
  \begin{itemize}
    \item Simple to compute, simple to explain to stakeholders.
    \item ``What fraction of cheese batches did we classify correctly?''
    \item Default metric in many tutorials.
  \end{itemize}
  \vfill
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.metrics import accuracy_score
accuracy_score(y_true, y_pred)
\end{lstlisting}}
  \vfill
  \footnotesize Sounds reasonable. Let's see when it stops being useful. \emph{Next frame: we'll break this down into the four count types ($TP$, $FP$, $FN$, $TN$) that every other metric is built from.}
\end{frame}

\begin{frame}{Predict-first: where accuracy breaks}
  Your cheese factory has 1000 daily batches. 970 are clean, 30 are contaminated. You ship a model that predicts ``clean'' for \emph{every single batch} (no actual model - just the constant).
  \vfill
  \pause
  \textbf{Predict:}
  \begin{itemize}
    \item What's the accuracy of this ``model''?
    \item What's its recall on contaminated batches?
    \item Would you ship it?
  \end{itemize}
  \pause
  \vfill
  \begin{itemize}
    \item \textbf{Accuracy: $970 / 1000 = 97\%$.} Looks excellent.
    \item \textbf{Recall on positives: $0\%$.} Catches none of the contaminated batches.
    \item \textbf{Would you ship?} Customers get sick. Lawsuits. Recall costs millions.
  \end{itemize}
  \vfill
  \fcolorbox{armred}{armred!8}{\parbox{0.96\textwidth}{\small
    \textbf{The fundamental problem with accuracy:} it weights every error equally. When classes are imbalanced, the rare class becomes invisible. \emph{The metric we need must focus on the positive class.}
  }}
\end{frame}

% ================================================================
\section{Precision, recall, F1}

\begin{frame}{The confusion matrix}
  All classification metrics derive from four counts. Lay them on a $2 \times 2$:
  \vfill
  \begin{center}
    \begin{tikzpicture}[every node/.style={font=\small}]
      \node at (1.5, 2.0) {\textbf{Pred.\ Positive}};
      \node at (4.5, 2.0) {\textbf{Pred.\ Negative}};
      \node at (-1.5, 1.0) {\textbf{Actual Pos.}};
      \node at (-1.5, -0.5) {\textbf{Actual Neg.}};
      \draw[armred, thick, fill=armred!10] (0, 0.4) rectangle (3, 1.6);
      \node[armred] at (1.5, 1.1) {\textbf{TP}};
      \node[armred, font=\scriptsize] at (1.5, 0.7) {bad cheese flagged};
      \draw[armorange, thick, fill=armorange!10] (3, 0.4) rectangle (6, 1.6);
      \node[armorange!80!black] at (4.5, 1.1) {\textbf{FN}};
      \node[armorange!80!black, font=\scriptsize] at (4.5, 0.7) {bad cheese missed!};
      \draw[armorange, thick, fill=armorange!10] (0, -1.1) rectangle (3, 0.1);
      \node[armorange!80!black] at (1.5, -0.4) {\textbf{FP}};
      \node[armorange!80!black, font=\scriptsize] at (1.5, -0.8) {wasted inspection};
      \draw[paramgreen, thick, fill=paramgreen!10] (3, -1.1) rectangle (6, 0.1);
      \node[paramgreen] at (4.5, -0.4) {\textbf{TN}};
      \node[paramgreen, font=\scriptsize] at (4.5, -0.8) {clean cheese passed};
    \end{tikzpicture}
  \end{center}
  \vfill
  \footnotesize \textbf{Asymmetry of cost in the cheese factory:}
  \begin{itemize}\footnotesize
    \item \textbf{FN (orange, top-right):} contaminated cheese reaches customers $\to$ recall, illness, lawsuit. \emph{The expensive mistake.}
    \item \textbf{FP (orange, bottom-left):} clean batch flagged for extra inspection $\to$ wasted hour of staff time. \emph{Annoying but cheap.}
    \item \textbf{TP (red):} caught a contaminated batch. \emph{The win.}
    \item \textbf{TN (green):} clean batch passed normally. \emph{The default.}
  \end{itemize}
  \vskip 2pt
  \scriptsize \emph{Heads-up:} sklearn's \texttt{confusion\_matrix} returns rows/cols as $[TN, FP; FN, TP]$ (negative class first) by default. Pass \texttt{labels=[1,0]} to flip into the positive-first layout shown above.
\end{frame}

\begin{frame}[fragile]{Precision and recall}
  \begin{columns}[T]
    \column{0.5\textwidth}
    \textbf{Recall} (also: sensitivity, TPR):
    \[
      \text{Recall} = \frac{TP}{TP + FN}
    \]
    \footnotesize ``Of all bad cheese batches, how many did the model catch?'' \emph{What fraction of contaminated cheese reaches customers? $1 - $ recall.}
    \vskip 6pt
    \textbf{For the cheese factory:} \emph{recall is the safety metric}. Every percentage point of recall translates to fewer sick customers.
    \column{0.5\textwidth}
    \textbf{Precision} (also: PPV):
    \[
      \text{Precision} = \frac{TP}{TP + FP}
    \]
    \footnotesize ``Of all batches the model flagged, how many were actually bad?'' \emph{What fraction of inspections were worthwhile?}
    \vskip 6pt
    \textbf{For the cheese factory:} \emph{precision is the efficiency metric}. Low precision $=$ inspectors chasing false alarms.
  \end{columns}
  \vfill
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.metrics import precision_score, recall_score
precision_score(y_true, y_pred)    # of flagged, how many real
recall_score(y_true, y_pred)       # of real, how many caught
\end{lstlisting}}
  \vfill
  \footnotesize \textbf{The tension:} flagging more batches raises recall but tanks precision. Flagging fewer raises precision but tanks recall.
\end{frame}

\begin{frame}[fragile]{F1: when you need a single number}
  Sometimes you need one metric instead of two. \textbf{F1 score} is the harmonic mean of precision and recall:
  \[
    F_1 = 2 \cdot \frac{\text{Precision} \cdot \text{Recall}}{\text{Precision} + \text{Recall}}
  \]
  \vfill
  \textbf{Why harmonic, not arithmetic?} Harmonic mean is dominated by the smaller value:
  \begin{center}\small
    \begin{tabular}{lccc}
      Precision & Recall & Arithmetic mean & Harmonic mean (F1) \\ \hline
      1.0 & 0.0 & 0.5 & 0.0 \\
      0.8 & 0.8 & 0.8 & 0.8 \\
      0.5 & 0.9 & 0.7 & 0.64 \\
      0.9 & 0.5 & 0.7 & 0.64
    \end{tabular}
  \end{center}
  \vfill
  Arithmetic mean lets a 1.0 in one column hide a 0.0 in the other. Harmonic mean refuses. F1 catches the ``cheating with extreme threshold'' that single metrics permit.
  \vfill
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.metrics import f1_score, fbeta_score
f1_score(y_true, y_pred)
fbeta_score(y_true, y_pred, beta=2)   # recall-leaning
\end{lstlisting}}
  \vfill
  \scriptsize \textbf{F$_\beta$ generalization:} $F_\beta = (1 + \beta^2) \cdot \dfrac{P \cdot R}{\beta^2 \cdot P + R}$. Higher $\beta$ weights recall more; $\beta = 2$ ($F_2$) fits the cheese factory better than $F_1$ because missing bad cheese is much worse than over-flagging clean.
\end{frame}

\begin{frame}{Worked example: cheese batch inspection}
  50 cheese batches inspected. 10 actually bad. Model flags 13.
  \vfill
  \begin{center}\small
    \begin{tabular}{lccc}
       & \textbf{Pred Bad} & \textbf{Pred Good} & \textbf{Total} \\ \hline
      Actual Bad (10) & TP $= 8$ & FN $= 2$ & 10 \\
      Actual Good (40) & FP $= 5$ & TN $= 35$ & 40 \\ \hline
      Total & 13 & 37 & 50
    \end{tabular}
  \end{center}
  \vfill
  \begin{center}\small
    \begin{tabular}{lll}
      \textbf{Metric} & \textbf{Calculation} & \textbf{Value} \\ \hline
      Accuracy & $(8 + 35) / 50$ & $86\%$ \\
      Recall & $8 / (8 + 2)$ & $80\%$ \\
      Precision & $8 / (8 + 5)$ & $62\%$ \\
      F1 & $2 \cdot 0.62 \cdot 0.80 / (0.62 + 0.80)$ & $70\%$ \\
    \end{tabular}
  \end{center}
  \vfill
  \footnotesize \textbf{What each metric says:}
  \begin{itemize}\footnotesize
    \item Accuracy: ``86\%, looks good.'' Misleading - hides the 2 bad batches that got through.
    \item Recall: ``Caught 80\% of contaminated batches.'' Operationally useful.
    \item Precision: ``Only 62\% of flagged batches were actually bad.'' 38\% wasted inspection effort.
    \item F1: ``70\%'' - a single number that punishes the imbalance between P and R.
  \end{itemize}
\end{frame}

% ================================================================
\section{Threshold dependence and the ROC fix}

\begin{frame}[fragile]{Where precision and recall come from: thresholds}
  \emph{Remember the precision-recall tension from the previous frame?} ``Flag more $\to$ recall up, precision down.'' \textbf{That tension IS the threshold knob.}
  \vfill
  Most classifiers don't output labels - they output \emph{scores} or \emph{probabilities}:
  \[
    \text{predict positive} \iff \text{score}(\xv) > c
  \]
  The \textbf{threshold $c$} is a knob we choose. Different $c$ $\Rightarrow$ different $(TP, FP, FN, TN)$ $\Rightarrow$ different P/R.
  \vfill
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
scores = model.predict_proba(X)[:, 1]    # probability of positive class
y_pred = (scores > 0.5).astype(int)       # default threshold
\end{lstlisting}}
  \vfill
  \begin{itemize}\small
    \item \textbf{Lower threshold} ($c \to 0$): flag more batches as bad $\to$ catch more (higher recall) but include more false alarms (lower precision).
    \item \textbf{Higher threshold} ($c \to 1$): flag fewer batches $\to$ fewer false alarms (higher precision) but miss more bad cheese (lower recall).
    \item \textbf{Default $c = 0.5$ is just one point on a curve.} Often not the right point.
  \end{itemize}
\end{frame}

\begin{frame}{Predict-first: P and R at extreme thresholds}
  Your model outputs scores. Consider two extreme threshold choices.
  \vfill
  \pause
  \textbf{Predict:} for each threshold, what are P and R?
  \begin{itemize}
    \item $c = 0.0$: predict positive for \emph{every batch}.
    \item $c = 1.0$: predict negative for \emph{every batch}.
  \end{itemize}
  \pause
  \vfill
  Reveal:
  \begin{itemize}\small
    \item \textbf{$c = 0$:} every batch predicted positive. $TP$ = all real positives, $FN = 0$. So \textbf{recall = 100\%}. $FP$ = all real negatives. \textbf{Precision = base rate} ($\sim 3\%$ for cheese).
    \item \textbf{$c = 1$:} every batch predicted negative. $TP = 0$, $FN$ = all positives. \textbf{Recall = 0\%}. Precision $= 0 / 0$ - sklearn returns $0$ by convention.
  \end{itemize}
  \vfill
  \fcolorbox{armred}{armred!8}{\parbox{0.96\textwidth}{\small
    \textbf{The problem with P/R as single metrics:} each one can be trivially gamed by an extreme threshold. F1 prevents the gaming but \emph{still depends on which threshold you picked}. We need a metric that summarizes performance \emph{across all thresholds at once}.
  }}
\end{frame}

\begin{frame}{The ROC curve}
  Two new names for old ratios: \textbf{TPR} $= \text{recall} = TP/(TP+FN)$ and \textbf{FPR} $= FP/(FP+TN)$ -- \emph{``of all clean batches, what fraction did we false-alarm on?''} (also called the false alarm rate).
  \vskip 2pt
  Sweep the threshold from 1 (no positives) down to 0 (all positives). At each step, record the (FPR, TPR) point. The trail is the \textbf{ROC curve}.
  \vfill
  \begin{columns}[T]
    \column{0.5\textwidth}
    \footnotesize
    \textbf{Algorithm.}
    \begin{enumerate}\footnotesize
      \item Rank predictions by score, highest first.
      \item Start at $c = 1$ $\Rightarrow$ everything predicted negative $\Rightarrow$ (FPR, TPR) $= (0, 0)$.
      \item Lower threshold so the top-ranked point becomes positive.
      \item If real positive: move TPR up by $1/n_+$. If real negative: move FPR right by $1/n_-$.
      \item Continue until $c = 0$ $\Rightarrow$ (FPR, TPR) $= (1, 1)$.
    \end{enumerate}
    \column{0.5\textwidth}
    \begin{center}
      \begin{tikzpicture}
        \begin{axis}[width=\textwidth, height=4.5cm,
                     xlabel={FPR}, ylabel={TPR},
                     xmin=0, xmax=1, ymin=0, ymax=1,
                     xlabel style={font=\scriptsize}, ylabel style={font=\scriptsize},
                     xticklabel style={font=\tiny}, yticklabel style={font=\tiny},
                     axis line style={gray!60},
                     grid=major, grid style={dashed,gray!30}]
          \draw[gray!50, dashed] (0,0) -- (1,1);
          \addplot[armred, very thick, mark=*, mark size=1pt]
            coordinates {(0,0) (0,0.2) (0,0.4) (0,0.6)
                         (0.14,0.6) (0.14,0.8) (0.29,0.8) (0.29,1.0)
                         (0.43,1.0) (0.57,1.0) (0.71,1.0) (0.86,1.0) (1.0,1.0)};
        \end{axis}
      \end{tikzpicture}\\
      \scriptsize ROC curve from 12-batch example. Each step is one ranked prediction crossing the threshold.
    \end{center}
  \end{columns}
  \vfill
  \footnotesize \textbf{Reading the curve:} top-left corner $=$ perfect (TPR $= 1$ at FPR $= 0$). Diagonal $=$ random guessing. Below diagonal $=$ worse than random (flip labels).
\end{frame}

\begin{frame}[fragile]{ROC AUC}
  Summarize the whole ROC curve as one number: the \textbf{area under the curve}.
  \vfill
  \begin{columns}[T]
    \column{0.5\textwidth}
    \footnotesize
    \begin{itemize}\footnotesize
      \item AUC $= 1.0$: perfect classifier (curve hits top-left).
      \item AUC $= 0.5$: random (curve $=$ diagonal).
      \item AUC $= 0.0$: perfectly wrong (flip labels for AUC $= 1$).
    \end{itemize}
    \vskip 6pt
    \textbf{Probabilistic interpretation.} ROC AUC $=$ probability that a random positive scores higher than a random negative. \emph{Equivalent to the Mann-Whitney $U$ / Wilcoxon rank-sum statistic -- a real probability, not just an area.}
    \vskip 6pt
    \textbf{Threshold-free.} You compare models on AUC \emph{without picking a threshold}. The metric summarizes the score's ranking quality.
    \column{0.5\textwidth}
    \begin{center}
      \begin{tikzpicture}
        \begin{axis}[width=\textwidth, height=4.2cm,
                     xlabel={FPR}, ylabel={TPR},
                     xmin=0, xmax=1, ymin=0, ymax=1,
                     xlabel style={font=\scriptsize}, ylabel style={font=\scriptsize},
                     xticklabel style={font=\tiny}, yticklabel style={font=\tiny},
                     axis line style={gray!60}]
          \addplot[name path=curve, armred, very thick, smooth]
            coordinates {(0,0) (0.05,0.4) (0.15,0.7) (0.3,0.85) (0.5,0.93) (0.75,0.98) (1.0,1.0)};
          \addplot[name path=axis, draw=none] coordinates {(0,0) (1.0,0)};
          \addplot[fill=armred!20, opacity=0.6] fill between[of=curve and axis];
          \draw[gray!50, dashed] (axis cs:0,0) -- (axis cs:1,1);
          \node[font=\small, armred] at (axis cs:0.5, 0.5) {AUC};
        \end{axis}
      \end{tikzpicture}\\
      \scriptsize Shaded area $=$ AUC. Higher $=$ better.
    \end{center}
  \end{columns}
  \vfill
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.metrics import roc_auc_score
roc_auc_score(y_true, y_scores)    # NOT y_pred - pass scores/probabilities!
\end{lstlisting}}
  \vfill
  \footnotesize \textbf{Common footgun:} passing predicted labels instead of probabilities collapses AUC to look like accuracy. Pass \texttt{model.predict\_proba(X)[:, 1]}.
\end{frame}

% ================================================================
\section{When ROC fails: PR and lift}

\begin{frame}{Predict-first: ROC's blind spot on imbalanced data}
  Two models on the same imbalanced cheese data: 100 bad batches, 10{,}000 clean batches.
  \vfill
  \begin{center}\small
    \begin{tabular}{lcccc}
      & \textbf{TP} & \textbf{FP} & \textbf{FN} & \textbf{TN} \\ \hline
      Model A & 91 & 10 & 9 & 9{,}990 \\
      Model B & 91 & 100 & 9 & 9{,}900 \\
    \end{tabular}
  \end{center}
  \vfill
  \pause
  \textbf{Predict:} for each model, compute
  \begin{itemize}\small
    \item Recall (TPR)?
    \item FPR?
    \item Precision (PPV)?
    \item Which model would you ship?
  \end{itemize}
  \pause
  \vfill
  \begin{center}\small
    \begin{tabular}{lccc}
      & \textbf{Recall} & \textbf{FPR} & \textbf{Precision} \\ \hline
      Model A & 91\% & $10/10{,}000 = \mathbf{0.10\%}$ & $91/101 = \mathbf{90\%}$ \\
      Model B & 91\% & $100/10{,}000 = \mathbf{1.0\%}$ & $91/191 = \mathbf{48\%}$ \\
    \end{tabular}
  \end{center}
  \vfill
  \fcolorbox{armred}{armred!8}{\parbox{0.96\textwidth}{\small
    \textbf{ROC's blind spot:} FPR moved from 0.10\% to 1.0\% - looks like a 0.9 percentage point difference. \emph{Tiny in ROC space.} Meanwhile, precision crashed from 90\% to 48\% - half of B's inspections are now wasted. \emph{On imbalanced data, the large negative class dominates the FPR denominator and hides huge operational differences.}
  }}
\end{frame}

\begin{frame}{The PR curve}
  Same sweep over thresholds, but plot \textbf{precision vs.\ recall} instead of TPR vs.\ FPR.
  \vfill
  \begin{columns}[T]
    \column{0.5\textwidth}
    \footnotesize
    \textbf{Reading the PR curve:}
    \begin{itemize}\footnotesize
      \item Top-right corner $=$ perfect (precision $= 1$, recall $= 1$).
      \item Top-left = high precision, low recall (strict threshold, few but high-quality alarms).
      \item Bottom-right = low precision, high recall (loose threshold, many alarms with many false).
      \item Baseline: a constant ``predict positive'' classifier has precision $=$ base rate, recall $= 1$.
    \end{itemize}
    \vskip 6pt
    \textbf{Doesn't depend on TN.} The TN count never enters precision or recall. \emph{This is exactly what makes PR informative on imbalanced data.}
    \vskip 4pt
    \textbf{Real PR curves zig-zag} (precision can rise as recall increases). The smooth curve here is a sketch; sklearn's \texttt{average\_precision\_score} handles the steps via a discrete sum.
    \column{0.5\textwidth}
    \begin{center}
      \begin{tikzpicture}
        \begin{axis}[width=\textwidth, height=4.2cm,
                     xlabel={Recall}, ylabel={Precision},
                     xmin=0, xmax=1, ymin=0, ymax=1,
                     xlabel style={font=\scriptsize}, ylabel style={font=\scriptsize},
                     xticklabel style={font=\tiny}, yticklabel style={font=\tiny},
                     axis line style={gray!60}, grid=major, grid style={dashed,gray!30}]
          \addplot[armred, very thick, smooth, mark=*, mark size=1pt]
            coordinates {(0,1.0) (0.1,0.95) (0.25,0.92) (0.5,0.85) (0.7,0.75) (0.85,0.6) (0.95,0.4) (1.0,0.05)};
          \draw[gray!50, dashed] (axis cs:0,0.05) -- (axis cs:1,0.05);
          \node[font=\tiny, gray] at (axis cs:0.5, 0.10) {base rate};
        \end{axis}
      \end{tikzpicture}\\
      \scriptsize PR curve. Top-right $=$ best. Baseline near the bottom $=$ class imbalance.
    \end{center}
  \end{columns}
\end{frame}

\begin{frame}[fragile]{PR AUC (Average Precision)}
  Summarize the PR curve as a single number: \textbf{average precision} (the area under the PR curve).
  \[
    \text{AP} = \sum_n (R_n - R_{n-1}) \cdot P_n
  \]
  where $n$ indexes the \emph{distinct recall thresholds} produced by sorting predictions by score, and $(P_n, R_n)$ are the precision and recall at threshold $n$.
  \vfill
  \begin{itemize}
    \item AP $= 1.0$: perfect ranking.
    \item AP $=$ \textbf{base rate}: random ranking. \emph{Not 0.5!}
    \item Higher AP $=$ better ranking, weighted toward high-precision regions.
  \end{itemize}
  \vfill
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.metrics import average_precision_score
average_precision_score(y_true, y_scores)
\end{lstlisting}}
  \vfill
  \fcolorbox{armred}{armred!8}{\parbox{0.96\textwidth}{\small
    \textbf{Sensitive to imbalance, by design.} The model A vs.\ B example: ROC AUC barely changes; PR AUC drops dramatically because precision drops dramatically. \emph{For imbalanced classification, PR AUC is the metric that matches operational reality.}
  }}
\end{frame}

\begin{frame}{Lift and gain charts: the ``top-$k$\%'' question}
  Operational constraint: the cheese factory has inspection capacity for the top 20\% of suspicious batches per day. \emph{Which batches should we inspect, and how much better than random?}
  \vfill
  \begin{columns}[T]
    \column{0.55\textwidth}
    \footnotesize
    \textbf{Cumulative gain.} Rank batches by score (highest first). $x$-axis: fraction inspected. $y$-axis: fraction of bad batches caught. Random $=$ diagonal; perfect $=$ vertical climb to $100\%$ by the base rate.
    \vskip 8pt
    \textbf{Lift.} $\text{lift}_k = \dfrac{\text{\% bad caught in top }k\%}{k\%}$
    \vskip 4pt
    Random $=$ horizontal line at $1.0$. Lift of $5$ at the top decile $=$ the model catches 5x more contaminated batches than random at the same inspection budget.
    \vskip 8pt
    \textbf{Why this matters.} ROC and PR AUC summarize the whole curve; lift answers the question that production actually asks: \emph{given my fixed budget, how much better than random?}
    \column{0.45\textwidth}
    \begin{center}
      \begin{tikzpicture}
        \begin{axis}[width=\textwidth, height=4cm,
                     xlabel={fraction of batches inspected}, ylabel={\% bad caught},
                     xmin=0, xmax=1, ymin=0, ymax=1.05,
                     xlabel style={font=\tiny}, ylabel style={font=\tiny},
                     xticklabel style={font=\tiny}, yticklabel style={font=\tiny},
                     axis line style={gray!60},
                     legend pos=south east, legend style={font=\tiny}]
          \addplot[gray!60, dashed] coordinates {(0,0) (1,1)};
          \addlegendentry{random}
          \addplot[paramgreen, very thick] coordinates {(0,0) (0.03,1.0) (1,1)};
          \addlegendentry{perfect}
          \addplot[armred, very thick] coordinates {(0,0) (0.05,0.4) (0.1,0.65) (0.2,0.85) (0.3,0.93) (0.5,0.98) (1,1)};
          \addlegendentry{our model}
          \draw[armorange, very thick, dashed] (axis cs:0.2, 0) -- (axis cs:0.2, 0.85);
          \node[font=\tiny, armorange] at (axis cs:0.25, 0.4) {top 20\%};
        \end{axis}
      \end{tikzpicture}\\
      \scriptsize At top 20\% (orange line) we catch $\sim 85\%$ of bad batches. Lift $\approx 85/20 = 4.25$.
    \end{center}
  \end{columns}
\end{frame}

\begin{frame}{ROC vs.\ PR side-by-side on imbalanced data}
  Same imbalanced dataset (1\% positive rate), same two models. Two views:
  \vfill
  \begin{center}
    \begin{tikzpicture}
      \begin{groupplot}[
        group style={group size=2 by 1, horizontal sep=18pt},
        width=0.42\textwidth, height=4cm,
        xmin=0, xmax=1, ymin=0, ymax=1,
        xtick={0,0.5,1}, ytick={0,0.5,1},
        xticklabel style={font=\tiny}, yticklabel style={font=\tiny},
        xlabel style={font=\scriptsize}, ylabel style={font=\scriptsize},
        axis line style={gray!60}, grid=major, grid style={dashed,gray!30},
        title style={font=\footnotesize}, legend style={font=\tiny},
      ]
        \nextgroupplot[title={ROC: ``both look great''}, xlabel={FPR}, ylabel={TPR}, legend pos=south east]
          \draw[gray!50, dashed] (axis cs:0,0) -- (axis cs:1,1);
          \addplot[armblue, very thick] coordinates {(0,0) (0.001,0.5) (0.005,0.85) (0.02,0.93) (0.1,0.98) (1,1)};
          \addlegendentry{Model A}
          \addplot[armred, very thick] coordinates {(0,0) (0.005,0.5) (0.02,0.85) (0.08,0.93) (0.2,0.98) (1,1)};
          \addlegendentry{Model B}
        \nextgroupplot[title={PR: ``actually broken''}, xlabel={Recall}, ylabel={Precision}, legend pos=south west]
          \addplot[armblue, very thick] coordinates {(0,1.0) (0.2,0.95) (0.5,0.85) (0.8,0.7) (0.95,0.4) (1.0,0.01)};
          \addlegendentry{Model A}
          \addplot[armred, very thick] coordinates {(0,1.0) (0.2,0.5) (0.5,0.3) (0.8,0.15) (0.95,0.05) (1.0,0.01)};
          \addlegendentry{Model B}
          \draw[gray!50, dashed] (axis cs:0,0.01) -- (axis cs:1,0.01);
      \end{groupplot}
    \end{tikzpicture}
  \end{center}
  \vfill
  \fcolorbox{armred}{armred!8}{\parbox{0.96\textwidth}{\small
    \textbf{Same models, two stories.} ROC: both close to top-left, AUCs $\approx 0.99$. PR: model A holds precision; model B collapses to near-baseline. On imbalanced data, \emph{ROC overstates and PR clarifies}. Reach for PR (or report both) whenever the positive class is rare.
  }}
\end{frame}

% ================================================================
\section{Practical: choosing, multi-class, tuning}

\begin{frame}{Which metric for which problem?}
  \begin{center}\small
    \begin{tabular}{p{4.5cm} p{4.5cm} p{4cm}}
      \textbf{Situation} & \textbf{Use} & \textbf{Why} \\ \hline
      Balanced classes, single threshold & F1 (or accuracy) & Single number; symmetric. \\
      Balanced classes, threshold-free & ROC AUC & Compares models on ranking quality. \\
      \textbf{Imbalanced + threshold-free} & PR AUC (average precision) & ROC overstates; PR matches operational reality. \\
      Bounded inspection budget & Lift at top-$k$\%, gain chart & Capacity is the real constraint. \\
      FN much worse than FP (cheese, fraud, medical screening) & Recall, with precision constraint & Asymmetric cost. \\
      FP much worse than FN (spam, churn intervention) & Precision, with recall constraint & Asymmetric cost, other direction. \\
      Single imbalance-robust number & MCC (Matthews CC) & Symmetric; $\in [-1, 1]$; $0 =$ random. \\
    \end{tabular}
  \end{center}
  \vfill
  \scriptsize \textbf{Default:} report 3-4 metrics. One threshold-free (ROC or PR) plus precision/recall at a chosen threshold. \emph{Calibration (Brier, log loss, ECE) -- separate deck.}
\end{frame}

\begin{frame}[fragile]{Multi-class metrics: macro vs.\ micro vs.\ weighted}
  With $K > 2$ classes, binary precision/recall don't apply directly. Compute one-vs-rest per class, then average. Three strategies:
  \vfill
  \begin{itemize}\small
    \item \textbf{Macro:} average per-class metric, \emph{equal weight per class}. Best when all classes matter equally regardless of size. ``How well does the model do on each cheese type?''
    \item \textbf{Micro:} pool all TP/FP/FN counts across classes, then compute. \emph{Equivalent to accuracy} for single-label classification. Best when overall correctness matters.
    \item \textbf{Weighted:} per-class metric weighted by class support (number of samples). Reflects the data distribution. \emph{Common default in sklearn reports.}
  \end{itemize}
  \vfill
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.metrics import f1_score, classification_report
f1_score(y_true, y_pred, average="macro")     # fair to small classes
f1_score(y_true, y_pred, average="micro")     # ~ accuracy
f1_score(y_true, y_pred, average="weighted")  # support-weighted

print(classification_report(y_true, y_pred))   # all three + per-class
\end{lstlisting}}
  \vfill
  \footnotesize \textbf{For multi-class ROC AUC:} use \texttt{roc\_auc\_score(y, scores, multi\_class="ovr", average="macro")}. Macro-OVR is the standard for ``how well does the model rank each class?''.
  \vskip 2pt
  \scriptsize \emph{Macro caveat:} when some classes have few samples (say $< 30$), per-class F1 is noisy and macro F1 inherits that noise. Always report \texttt{support} (number of samples per class) alongside.
\end{frame}

\begin{frame}[fragile]{Threshold tuning: the threshold is its own decision}
  Default threshold ($c = 0.5$) is just a starting point. The \emph{optimal} threshold depends on the cost asymmetry.

  \textbf{Cost-sensitive procedure.} Let $C_\text{FN}$ $=$ cost of missing a bad batch, $C_\text{FP}$ $=$ cost of inspecting a clean one. Then:
  \[
    c^* = \argmin_c \; C_\text{FN} \cdot FN(c) + C_\text{FP} \cdot FP(c)
  \]
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.metrics import precision_recall_curve, confusion_matrix
p, r, thresholds = precision_recall_curve(y_val, scores_val)
C_fn, C_fp = 1000, 10           # $1000 per miss, $10 per inspection
costs = []
for t in thresholds:
    pred = (scores_val > t).astype(int)
    tn, fp, fn, tp = confusion_matrix(y_val, pred).ravel()
    costs.append(C_fn * fn + C_fp * fp)
best_t = thresholds[np.argmin(costs)]
\end{lstlisting}}
  \scriptsize \fcolorbox{armred}{armred!8}{\parbox{0.97\linewidth}{\textbf{Tune on a held-out set}, not the data you measured the metric on -- otherwise the threshold is optimistically biased too.}}
\end{frame}

% ================================================================
\section{Wrap-up}

\begin{frame}{Decision flowchart}
  \begin{center}
    \begin{tikzpicture}[
      node distance=0.4cm and 0.7cm,
      box/.style={rectangle, draw=armblue, very thick, text width=3cm, align=center, font=\scriptsize, rounded corners=2pt, fill=armblue!5, minimum height=0.85cm},
      result/.style={rectangle, draw=paramgreen, very thick, text width=3cm, align=center, font=\scriptsize, rounded corners=2pt, fill=paramgreen!10, minimum height=0.85cm},
      arr/.style={->, >=stealth, thick, armblue},
    ]
      \node[box] (start) {Multi-class?};
      \node[result, right=of start] (mc) {macro/micro/weighted F1; classification\_report};
      \node[box, below=of start] (imb) {Imbalanced classes?};
      \node[result, right=of imb] (pr) {PR AUC + precision/recall at chosen threshold};
      \node[box, below=of imb] (cap) {Bounded inspection / contact budget?};
      \node[result, right=of cap] (lift) {Lift / gain at top-$k$\%};
      \node[box, below=of cap] (sym) {Symmetric cost, balanced data?};
      \node[result, right=of sym] (roc) {ROC AUC + F1};
      \node[result, below=of sym] (cost) {Accuracy + threshold tuned by cost};

      \draw[arr] (start) -- node[above, font=\tiny]{yes} (mc);
      \draw[arr] (start) -- node[left, font=\tiny]{no} (imb);
      \draw[arr] (imb) -- node[above, font=\tiny]{yes} (pr);
      \draw[arr] (imb) -- node[left, font=\tiny]{no} (cap);
      \draw[arr] (cap) -- node[above, font=\tiny]{yes} (lift);
      \draw[arr] (cap) -- node[left, font=\tiny]{no} (sym);
      \draw[arr] (sym) -- node[above, font=\tiny]{yes} (roc);
      \draw[arr] (sym) -- node[left, font=\tiny]{no} (cost);
    \end{tikzpicture}
  \end{center}
  \vfill
  \footnotesize \textbf{Always report multiple metrics.} A single number hides the trade-off; the trade-off is the story.
  \vskip 2pt
  \scriptsize \emph{Caveat:} real problems often hit multiple branches at once (multi-class AND imbalanced, bounded budget AND asymmetric cost). The tree picks a \emph{primary} metric; report secondaries too.
\end{frame}

\begin{frame}{Recap: the cascade}
  \textbf{The metric cascade -- each one fixes the previous one's problem:}
  \begin{enumerate}\small
    \item \textbf{Accuracy} - simple, fails on imbalanced data (97\% accuracy with 0\% recall).
    \item \textbf{Precision and recall} - focus on the positive class. \emph{But depend on the threshold.}
    \item \textbf{F1} - one number that punishes asymmetric P/R. \emph{Still depends on the threshold.}
    \item \textbf{ROC AUC} - threshold-free; ranking quality. \emph{Overstates on imbalanced data.}
    \item \textbf{PR AUC (average precision)} - threshold-free AND sensitive to imbalance. The right metric for rare-class problems.
    \item \textbf{Lift / gain} - the operational ``top-$k$\%'' question when capacity is bounded.
  \end{enumerate}
  \vskip 4pt
  \textbf{Practical machinery -- not part of the cascade, but you'll need them:}
  \begin{itemize}\small
    \item \textbf{Multi-class:} macro for fairness across classes, micro for overall accuracy, weighted for distribution-aware.
    \item \textbf{Threshold tuning:} given the cost asymmetry, the optimal threshold rarely equals 0.5.
  \end{itemize}
  \vskip 4pt
  \footnotesize \textbf{One-line takeaway:} report multiple metrics, always include one that matches your data's imbalance and your problem's cost structure.
\end{frame}

\begin{frame}{HW01 - cheese inspector practice}
  \scriptsize Estimated time: 90-120 minutes. Tasks 1-4 graded; bonuses optional.
  \begin{enumerate}\scriptsize
    \item Generate a synthetic cheese-batch dataset: $n = 1000$, 3\% contaminated (\texttt{make\_classification}, \texttt{weights=[0.97, 0.03]}). Stratified train/test split.
    \item Fit logistic regression. Compute accuracy, precision, recall, F1, ROC AUC, average precision (PR AUC) on test. Discuss which metrics agree and which disagree.
    \item Plot confusion matrix at threshold $0.5$, ROC curve, PR curve. Identify the operating point that maximizes F1.
    \item Plot cumulative gain and lift charts. At what top-$k$\% does the model catch 80\% of bad batches? Compute lift at top 20\%. \emph{Hint:} \texttt{np.searchsorted(cum\_gain, 0.80)}.
    \item \emph{Bonus 1:} change class balance to $1\%$ positives. Re-run all metrics. Which changed most? Tie back to frame 13.
    \item \emph{Bonus 2 (most instructive):} compare ROC AUC vs.\ PR AUC as imbalance sweeps from $50\%$ down to $0.5\%$. Plot both; show the gap grows.
    \item \emph{Bonus 3:} cost-sensitive threshold tuning. $C_\text{FN} = 1000$, $C_\text{FP} = 10$. Find $c^*$ minimizing total cost. Compare to F1-optimal threshold.
    \item \emph{Bonus 4:} multi-class (3 cheese types: good, off-flavor, contaminated). Compute macro/micro/weighted F1 and discuss which best reflects the factory's priorities.
  \end{enumerate}
\end{frame}

\end{document}

% =============================================================
% Provenance:
%   - Topic from plan_manual.md line 30 (classification metrics).
%   - Reference: ml/Chapter 2 Classification/PDFs/C2 02 Classification evaluation metrics.pdf
%   - User requested cascading narrative: accuracy -> P/R -> ROC -> PR
%   - Running example: cheese factory contamination detection
%   - Predict-first frames at 3 pivot points
%   - Includes lift/gain chart per user request
%   - Multi-class macro/micro/weighted frame included
%   - Dedicated threshold-tuning frame
% =============================================================
