\documentclass[aspectratio=169]{beamer}
\input{../preamble}

\usepgfplotslibrary{groupplots, fillbetween}
\usepackage{listings}

\lstset{
  basicstyle=\ttfamily\footnotesize,
  keywordstyle=\color{armblue}\bfseries,
  commentstyle=\color{paramgreen}\itshape,
  stringstyle=\color{armred},
  showstringspaces=false,
  language=Python,
  frame=single,
  framerule=0pt,
  backgroundcolor=\color{lightbg},
  numbers=none,
  breaklines=true,
  xleftmargin=4pt,
  xrightmargin=4pt,
}

\title{L01h: Feature Selection}
\subtitle{Which features matter, and how to be sure}
\date{}

\begin{document}

\begin{frame}{Outline}
  \tableofcontents
\end{frame}

% ================================================================
\section{Why and how}

\begin{frame}{Why select features?}
  Five reasons to drop features:
  \begin{enumerate}
    \item \textbf{Curse of dimensionality} (math module 30). Volume of $p$-dim space grows exponentially. With $n$ fixed, samples become sparse.
    \item \textbf{Multicollinearity} (L01b). Correlated features make linear coefficients unstable; $\Xmat^\top \Xmat$ becomes ill-conditioned.
    \item \textbf{Overfitting risk.} More features $\Rightarrow$ more flexibility $\Rightarrow$ higher variance. Regularization helps but doesn't eliminate.
    \item \textbf{Compute and storage.} Linear-in-$p$ at inference, polynomial during fit.
    \item \textbf{Interpretability.} ``We use 8 features'' defends better than ``we use 800''. Critical for healthcare, credit, audit.
  \end{enumerate}
  \vfill
  \textbf{When NOT to select.} With heavy L1/L2 regularization, the model effectively selects \emph{soft}. Sometimes deleting features hurts more than it helps - the dropped features may carry small but real signal.
\end{frame}

\begin{frame}{Predict-first: does adding features always help?}
  You have $n = 200$ samples. You add features one at a time, fitting a Ridge model after each addition. Each new feature is a real signal (correlated with $y$) plus some noise.
  \vfill
  \pause
  \textbf{Predict:} as you keep adding features, what happens to
  \begin{itemize}
    \item training error?
    \item test error?
  \end{itemize}
  \pause
  \vfill
  \begin{columns}[T]
    \column{0.5\textwidth}
    \footnotesize
    Training error \textbf{always falls} - more features means more flexibility to fit the training set.
    \vskip 4pt
    Test error follows a \textbf{U-shape}. It falls while features carry signal, then rises as added features bring more variance than bias reduction. The minimum is the sweet spot.
    \vskip 4pt
    Feature selection is about finding that sweet spot.
    \column{0.5\textwidth}
    \begin{center}
      \begin{tikzpicture}
        \begin{axis}[width=\textwidth, height=4cm,
                     xlabel={features kept ($k$)},
                     ylabel={error}, xlabel style={font=\scriptsize},
                     ylabel style={font=\scriptsize}, ytick=\empty,
                     xtick={1,5,10,15,20}, xticklabel style={font=\tiny},
                     axis line style={gray!60}, ymin=0,
                     legend pos=north east, legend style={font=\tiny},
                     grid=major, grid style={dashed,gray!30}]
          \addplot[armblue, very thick, smooth] coordinates {
            (1,0.95) (3,0.7) (5,0.5) (7,0.36) (9,0.25)
            (11,0.18) (13,0.13) (15,0.09) (17,0.07) (20,0.04)};
          \addlegendentry{training}
          \addplot[armred, very thick, smooth] coordinates {
            (1,1.0) (3,0.7) (5,0.5) (7,0.42) (9,0.40)
            (11,0.45) (13,0.55) (15,0.70) (17,0.90) (20,1.25)};
          \addlegendentry{test}
          \addplot[armorange, only marks, mark=triangle*, mark size=4pt]
            coordinates {(9, 0.40)};
        \end{axis}
      \end{tikzpicture}\\
      \scriptsize Sweet spot at $k \approx 9$; more features hurt.
    \end{center}
  \end{columns}
\end{frame}

\begin{frame}{The three families of selection methods}
  \begin{center}\small
    \begin{tabular}{p{2.8cm} p{4.5cm} p{4.5cm}}
      \textbf{Family} & \textbf{How it works} & \textbf{Strengths / weaknesses} \\ \hline
      \textbf{Filter} & Score each feature \emph{independently} against the target. Pick top $k$. & Fast, model-agnostic. \emph{Blind to interactions.} \\[2pt]
      \textbf{Embedded} & Use a model that selects \emph{as it fits} (Lasso, tree importances). & Cheap, model-aware. Tied to the chosen model. \\[2pt]
      \textbf{Wrapper} & Search over feature subsets, evaluate each via CV. & Most accurate. Expensive: refits the model $\mathcal{O}(p)$-$\mathcal{O}(p^2)$ times. \\
    \end{tabular}
  \end{center}
  \vfill
  Cost goes up left-to-right. Accuracy (for the model you'll deploy) usually does too.
  \vfill
  \textbf{Typical workflow:}
  \begin{enumerate}\small
    \item Filter from $p = 10{,}000$ down to $p = 500$ (drop near-constants, very low MI).
    \item Embedded selection (Lasso or tree importance) on the 500 to get 50-100 candidates.
    \item Wrapper (RFE-CV or Boruta) on the 50-100 to get the final 10-20.
  \end{enumerate}
\end{frame}

% ================================================================
\section{Filter methods}

\begin{frame}[fragile]{Variance threshold and univariate filters}
  \begin{columns}[T]
    \column{0.55\textwidth}
    \footnotesize
    \textbf{Variance threshold.} Drop near-constant features. Quick win - constant features carry no signal.
    \vskip 4pt
    \textbf{Univariate statistical tests.} Score each feature by its individual association with $y$:
    \begin{itemize}\footnotesize
      \item Regression: \texttt{f\_regression} (linear), \texttt{mutual\_info\_regression} (any monotonic).
      \item Classification: \texttt{f\_classif} (ANOVA), \texttt{chi2} (categorical), \texttt{mutual\_info\_classif}.
    \end{itemize}
    \vskip 4pt
    Pick top $k$ via \texttt{SelectKBest} or percentile via \texttt{SelectPercentile}.
    \column{0.45\textwidth}
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.feature_selection import (
    VarianceThreshold,
    SelectKBest,
    f_regression,
    mutual_info_regression)

# Drop near-constants
X = VarianceThreshold(0.01).fit_transform(X)

# Top-20 by linear correlation
sel = SelectKBest(f_regression, k=20)
X_sel = sel.fit_transform(X, y)

# Top-20 by mutual information
sel_mi = SelectKBest(
    mutual_info_regression, k=20)
\end{lstlisting}}
  \end{columns}
\end{frame}

\begin{frame}{The independence trap}
  Filter methods score each feature \emph{alone}. They miss \textbf{interactions}.
  \vfill
  \begin{columns}[T]
    \column{0.45\textwidth}
    \footnotesize
    Consider 4 points: $(x_1, x_2, y)$
    \vskip 2pt
    \begin{center}\scriptsize
      \begin{tabular}{ccc}
        $x_1$ & $x_2$ & $y$ \\ \hline
        0 & 0 & 0 \\
        0 & 1 & 1 \\
        1 & 0 & 1 \\
        1 & 1 & 0
      \end{tabular}
    \end{center}
    \vskip 4pt
    Marginal correlations: $\text{corr}(x_1, y) = 0$, $\text{corr}(x_2, y) = 0$. Mutual information $\approx 0$ per feature.
    \vskip 4pt
    Jointly: $y = x_1 \oplus x_2$. Perfect predictor.
    \column{0.55\textwidth}
    Filter score on this data: \emph{both features look useless}. A linear model fed only $x_1$ or only $x_2$ confirms it.
    \vfill
    \textbf{Lesson:} filter is a screen, not a decision. A feature that fails the univariate test may still matter in combination.
    \vfill
    \textbf{Workarounds:}
    \begin{itemize}\small
      \item Add hand-engineered interactions before filtering (L01g).
      \item Use a multivariate filter (next slide).
      \item Use embedded methods on a tree model, which discovers interactions.
    \end{itemize}
  \end{columns}
\end{frame}

\begin{frame}{Multivariate filters}
  Filters that consider feature \emph{combinations} instead of one-at-a-time.
  \vfill
  \begin{itemize}
    \item \textbf{mRMR} (Minimum Redundancy Maximum Relevance). At each step, pick the feature that maximizes (relevance to $y$) minus (average correlation with already-picked features). Catches non-redundant signal.
    \item \textbf{Cluster-and-represent.} Group features by mutual correlation (hierarchical clustering), pick one representative per cluster (the one with highest target correlation). Cheap, removes obvious redundancy.
    \item \textbf{HSIC-Lasso} (Hilbert-Schmidt Independence Criterion). Kernel-based, captures non-linear dependence.
  \end{itemize}
  \vfill
  \footnotesize Libraries: \texttt{pymrmr}, \texttt{mrmr-selection}, \texttt{scikit-feature}. Not as common as filter+embedded combinations, but useful when $p$ is large and many features are correlated (genomics, sensor data).
\end{frame}

% ================================================================
\section{Embedded methods}

\begin{frame}[fragile]{L1 / Lasso for sparsity}
  Recall Ridge (L01b) minimizes $\|\yv - \Xmat\thetav\|^2 + \lambda\,\|\thetav\|_2^2$. The L2 penalty shrinks coefficients toward zero but rarely \emph{to} zero.
  \vfill
  \textbf{Lasso} replaces L2 with L1:
  \[
    \thetah_{\text{Lasso}} = \argmin_{\thetav}\; \| \yv - \Xmat\thetav \|^2 + \lambda \, \| \thetav \|_1
  \]
  The L1 penalty has a non-smooth corner at $\theta_j = 0$, which \emph{forces} coefficients exactly to zero. Features with zero coefficient are selected out.
  \vfill
  \begin{columns}[T]
    \column{0.5\textwidth}
    \begin{center}
      \begin{tikzpicture}
        \begin{axis}[width=\textwidth, height=3.5cm,
                     xlabel={$\log \lambda$ (increasing penalty $\to$)},
                     ylabel={coefficient $\theta_j$},
                     xlabel style={font=\tiny}, ylabel style={font=\tiny},
                     xtick=\empty, ytick=\empty,
                     axis line style={gray!60}, ymin=-1.5, ymax=2.0,
                     grid=major, grid style={dashed,gray!30}]
          \draw[gray!50, dashed] (axis cs:-3, 0) -- (axis cs:3, 0);
          \addplot[armred, very thick, smooth] coordinates {(-3, 1.9) (-2, 1.6) (-1, 1.2) (0, 0.8) (1, 0.4) (2, 0.1) (3, 0)};
          \addplot[armblue, very thick, smooth] coordinates {(-3, 1.3) (-2, 1.0) (-1, 0.7) (0, 0.4) (1, 0.1) (1.5, 0) (3, 0)};
          \addplot[paramgreen, very thick, smooth] coordinates {(-3, -0.9) (-2, -0.6) (-1, -0.3) (-0.3, 0) (3, 0)};
          \addplot[armorange, very thick, smooth] coordinates {(-3, 0.5) (-2, 0.3) (-1, 0.1) (-0.5, 0) (3, 0)};
        \end{axis}
      \end{tikzpicture}\\
      \scriptsize Lasso path. Less-important features hit 0 first.
    \end{center}
    \column{0.5\textwidth}
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.linear_model import LassoCV
from sklearn.feature_selection \
    import SelectFromModel

lasso = LassoCV(cv=5,
                random_state=509).fit(X, y)
sel = SelectFromModel(lasso,
                      prefit=True)
X_sel = sel.transform(X)
# Selected feature indices
keep = sel.get_support(indices=True)
\end{lstlisting}}
  \end{columns}
\end{frame}

\begin{frame}[fragile]{Tree impurity importance}
  Tree models track how much each feature reduces the loss across all its splits.
  \vfill
  \begin{itemize}
    \item For \textbf{regression trees}: importance$_j$ $=$ sum of (variance reduction at each split using feature $j$), weighted by samples reaching that split.
    \item For \textbf{classification trees}: same idea with Gini decrease or entropy decrease.
    \item Random Forest / GBM average importance across all trees.
  \end{itemize}
  \vfill
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.ensemble import RandomForestRegressor
rf = RandomForestRegressor(n_estimators=300,
                           random_state=509).fit(X, y)
importances = pd.Series(rf.feature_importances_, index=X.columns)
importances.sort_values(ascending=False).head(10)
\end{lstlisting}}
  \vfill
  Cheap (free with the model), intuitive, used everywhere. \textbf{Also: misleading} on correlated features and high-cardinality categoricals - the next slide is the most important warning in this deck.
\end{frame}

\begin{frame}{Predict-first: impurity importance on correlated features}
  You have 5 features. Features $x_1$ and $x_2$ are nearly perfectly correlated (cor $\approx 0.99$) and both strongly predict $y$. Features $x_3, x_4, x_5$ are independent and predictive.
  \vfill
  \pause
  \textbf{Predict:} when you fit a Random Forest and read off impurity importances, what does $x_1$'s importance look like?
  \pause
  \vfill
  \begin{columns}[T]
    \column{0.5\textwidth}
    \footnotesize
    Each tree splits on one of $\{x_1, x_2\}$ - whichever the random feature subsample saw first. The total ``true'' importance is split roughly half-and-half.
    \vskip 4pt
    \textbf{A naive reader concludes:} ``$x_1$ and $x_2$ are about as important as $x_3, x_4, x_5$. Drop $x_1$ to simplify.''
    \vskip 4pt
    \textbf{What actually happened:} $x_1$ alone is highly predictive; the importance is just shared with its near-duplicate. Dropping $x_1$ leaves $x_2$ to do the same work; dropping both is a disaster.
    \column{0.5\textwidth}
    \begin{center}
      \begin{tikzpicture}
        \begin{axis}[width=\textwidth, height=4cm,
                     xbar, xmin=0, xmax=0.45,
                     symbolic y coords={x5, x4, x3, x2, x1},
                     ytick=data,
                     xlabel={impurity importance},
                     xlabel style={font=\tiny},
                     yticklabel style={font=\scriptsize},
                     xticklabel style={font=\tiny},
                     nodes near coords, nodes near coords style={font=\tiny},
                     enlarge y limits=0.1,
                     axis line style={gray!60}]
          \addplot[fill=armred!50, draw=armred] coordinates {
            (0.18,x1) (0.18,x2) (0.20,x3) (0.21,x4) (0.23,x5)};
        \end{axis}
      \end{tikzpicture}\\
      \scriptsize $x_1$ and $x_2$ split the importance they jointly carry.
    \end{center}
  \end{columns}
\end{frame}

\begin{frame}[fragile]{Permutation importance: the more reliable alternative}
  \textbf{Idea.} Shuffle one column at a time; measure how much the model's score drops on a held-out set. Big drop $\Rightarrow$ feature mattered.
  \vfill
  \begin{columns}[T]
    \column{0.55\textwidth}
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.inspection import (
    permutation_importance)

model = RandomForestRegressor().fit(X_tr, y_tr)
res = permutation_importance(
    model, X_val, y_val,
    n_repeats=30,
    random_state=509)
# res.importances_mean, res.importances_std
\end{lstlisting}}
    \footnotesize Advantages:
    \begin{itemize}\footnotesize
      \item Model-agnostic - works for any \texttt{score} method.
      \item Reflects \emph{validation} performance, not training.
      \item Multiple shuffles give error bars.
    \end{itemize}
    \column{0.45\textwidth}
    \begin{center}
      \begin{tikzpicture}
        \begin{axis}[width=\textwidth, height=4cm,
                     xbar, xmin=0, xmax=0.4,
                     symbolic y coords={x5, x4, x3, x2, x1},
                     ytick=data,
                     xlabel={score drop after shuffle},
                     xlabel style={font=\tiny},
                     yticklabel style={font=\scriptsize},
                     xticklabel style={font=\tiny},
                     enlarge y limits=0.1,
                     axis line style={gray!60}]
          \addplot[fill=armblue!50, draw=armblue]
            coordinates {
              (0.32,x1) (0.30,x2) (0.15,x3) (0.10,x4) (0.02,x5)};
        \end{axis}
      \end{tikzpicture}\\
      \scriptsize Correlated $x_1, x_2$ are \emph{both} marked important.
    \end{center}
  \end{columns}
  \vfill
  \footnotesize \textbf{Caveat:} permutation breaks correlations the model relied on. If features are highly correlated, permutation importance over-credits them. The cure is to use \emph{conditional} permutation or pair with SHAP.
\end{frame}

\begin{frame}[fragile]{SHAP: model-agnostic, theoretically grounded}
  SHAP (SHapley Additive exPlanations) gives each feature its \emph{Shapley value} from cooperative game theory: the average marginal contribution to the prediction across all possible feature subsets.
  \vfill
  \begin{itemize}\small
    \item Works on any model. Fast for trees via \texttt{TreeExplainer} (exact, polynomial-time).
    \item Local (per-prediction) and global (mean absolute SHAP across samples) importance.
    \item Modern industry default for interpretability.
    \item Handles correlated features more honestly than impurity importance.
  \end{itemize}
  \vfill
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
import shap
explainer = shap.TreeExplainer(model)
shap_values = explainer.shap_values(X_val)         # one per (sample, feature)
mean_abs_shap = np.abs(shap_values).mean(axis=0)   # global importance
shap.summary_plot(shap_values, X_val)              # built-in plot
\end{lstlisting}}
  \vfill
  \footnotesize \textbf{Caveat:} ``importance'' here means contribution to \emph{the model's prediction}, not to the true causal effect. A correlated nuisance feature can have high SHAP value if the model happens to use it.
\end{frame}

% ================================================================
\section{Wrapper methods}

\begin{frame}[fragile]{RFE and RFE-CV}
  \textbf{Recursive Feature Elimination.} Fit the model, drop the least-important feature, refit, repeat until $k$ features remain.
  \vfill
  \begin{columns}[T]
    \column{0.5\textwidth}
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.feature_selection \
    import RFE, RFECV
from sklearn.linear_model import Ridge

# Fixed k
rfe = RFE(Ridge(), n_features_to_select=10).fit(X, y)

# Auto k via CV (preferred)
rfecv = RFECV(
    Ridge(), cv=5,
    scoring="neg_mean_squared_error",
    min_features_to_select=2).fit(X, y)
print(rfecv.n_features_)
\end{lstlisting}}
    \column{0.5\textwidth}
    \begin{center}
      \begin{tikzpicture}
        \begin{axis}[width=\textwidth, height=4cm,
                     xlabel={features kept ($k$)}, ylabel={CV score},
                     xlabel style={font=\tiny}, ylabel style={font=\tiny},
                     xtick={1,5,10,15,20}, xticklabel style={font=\tiny},
                     ytick=\empty,
                     axis line style={gray!60}, grid=major,
                     grid style={dashed,gray!30}]
          \addplot[armblue, very thick, smooth, mark=*, mark size=1pt]
            coordinates {(1,0.3) (3,0.55) (5,0.72) (7,0.82) (9,0.87)
                         (11,0.87) (13,0.86) (15,0.83) (17,0.79) (20,0.74)};
          \addplot[armorange, only marks, mark=triangle*, mark size=4pt]
            coordinates {(9, 0.87)};
        \end{axis}
      \end{tikzpicture}\\
      \scriptsize Elbow at $k = 9$. RFECV picks this.
    \end{center}
  \end{columns}
  \vfill
  \footnotesize Cost: $\mathcal{O}(p)$ model fits per fold. Use a fast model (Ridge / LogReg) or pre-filter to $p < 100$.
\end{frame}

\begin{frame}[fragile]{Sequential forward / backward selection}
  Greedy alternatives to RFE. Slower but flexible across model classes.
  \vfill
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.feature_selection import SequentialFeatureSelector

# Forward: start empty, add the feature that most improves CV
sfs_fwd = SequentialFeatureSelector(
    Ridge(), n_features_to_select=10,
    direction="forward", cv=5).fit(X, y)

# Backward: start full, drop the feature whose removal hurts CV least
sfs_bwd = SequentialFeatureSelector(
    Ridge(), n_features_to_select=10,
    direction="backward", cv=5).fit(X, y)
\end{lstlisting}}
  \vfill
  \begin{itemize}\small
    \item \textbf{Forward} - fast when target $k \ll p$ (few features kept). Each step refits with one more feature.
    \item \textbf{Backward} - fast when target $k \approx p$ (few features dropped). Each step refits with one fewer.
    \item Both miss good \emph{combinations} that don't show in single-step additions. RFE often does better.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]{Boruta: the ``all-relevant'' approach}
  Most methods find a \emph{minimal} relevant set. Boruta finds \emph{all} features that beat random chance.
  \vfill
  \textbf{Algorithm.}
  \begin{enumerate}\small
    \item For each real feature, create a \emph{shadow}: a shuffled copy. Shadow features carry no real signal.
    \item Fit a Random Forest on real $+$ shadows.
    \item Compare each real feature's importance to the \emph{maximum} shadow importance.
    \item Repeat for many iterations. A feature that consistently beats the max shadow is ``confirmed'', one that consistently doesn't is ``rejected''.
  \end{enumerate}
  \vfill
  \begin{columns}[T]
    \column{0.5\textwidth}
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from boruta import BorutaPy
from sklearn.ensemble import \
    RandomForestRegressor

rf = RandomForestRegressor(
    n_estimators=200, random_state=509)
br = BorutaPy(rf, max_iter=100,
              random_state=509).fit(
    X.values, y.values)
print(X.columns[br.support_])
\end{lstlisting}}
    \column{0.5\textwidth}
    \begin{center}
      \begin{tikzpicture}
        \begin{axis}[width=\textwidth, height=4cm,
                     xbar, xmin=0, xmax=0.45,
                     symbolic y coords={x5, shmax, x4, x3, x2, x1},
                     ytick=data,
                     xlabel={importance}, xlabel style={font=\tiny},
                     yticklabel style={font=\scriptsize},
                     xticklabel style={font=\tiny},
                     enlarge y limits=0.15,
                     axis line style={gray!60}]
          \addplot[fill=armred!50, draw=armred] coordinates {
            (0.35,x1) (0.30,x2) (0.20,x3) (0.10,x4) (0.05,x5)};
          \addplot[fill=gray!60, draw=gray] coordinates {(0.12,shmax)};
        \end{axis}
      \end{tikzpicture}\\
      \scriptsize Real features above shmax (gray) are kept; below it are rejected.
    \end{center}
  \end{columns}
\end{frame}

% ================================================================
\section{Stability and pitfalls}

\begin{frame}[fragile]{Stability selection + how many features?}
  ``Run Lasso once'' gives one answer. ``Run Lasso many times on bootstrap samples'' tells you which features show up \emph{robustly}.
  \vfill
  \textbf{Stability selection (Meinshausen \& Buhlmann, 2010).}
  \begin{enumerate}\small
    \item Draw $B = 100$ bootstrap subsamples.
    \item Fit Lasso (with a range of $\lambda$s) on each.
    \item For each feature, record fraction of bootstraps where it was selected at all.
    \item Keep features with selection frequency $> 0.5$ (or any chosen threshold).
  \end{enumerate}
  \vfill
{\lstset{basicstyle=\ttfamily\scriptsize}
\begin{lstlisting}
from sklearn.utils import resample
from sklearn.linear_model import LassoCV
import numpy as np

selected_counts = np.zeros(X.shape[1])
for _ in range(100):
    X_bs, y_bs = resample(X, y, random_state=None)
    sel = LassoCV(cv=5).fit(X_bs, y_bs)
    selected_counts += (sel.coef_ != 0)
stability = selected_counts / 100
keep = stability > 0.5
\end{lstlisting}}
  \vfill
  \footnotesize \textbf{How many features?} Stability selection answers it automatically - $k =$ count of features above the threshold. Alternative: pick $k$ at the RFE-CV elbow.
\end{frame}

\begin{frame}{Pitfalls}
  \begin{enumerate}
    \item \textbf{Selection bias (the biggest one).} Doing feature selection on the \emph{full} dataset before splitting inflates CV scores. The chosen features were peeked at including test rows. Fix: nested CV with selection inside the inner loop, or do selection on train fold only.
    \item \textbf{Leakage via target encoding.} If features were target-encoded (L01c) without CV, selection just propagates the leakage. Fix the encoder first.
    \item \textbf{Model-dependence.} Features selected by Lasso may differ from those selected by tree importance. Selection results don't transfer across model classes; re-select if you switch models.
    \item \textbf{Correlated features.} Impurity importance splits credit; Lasso picks arbitrarily; RFE drops one randomly. None handles correlation cleanly. SHAP and permutation help.
    \item \textbf{Over-aggressive thresholds.} ``Keep features with importance $> 0.01$'' is arbitrary. Use CV elbow, stability frequency, or Boruta's shadow comparison instead.
    \item \textbf{Forgetting the baseline.} Always compare to ``keep all features.'' Selection that doesn't improve CV \emph{wasn't worth doing}.
  \end{enumerate}
\end{frame}

\begin{frame}{Decision flowchart}
  \begin{center}
    \begin{tikzpicture}[
      node distance=0.4cm and 0.7cm,
      box/.style={rectangle, draw=armblue, very thick, text width=3cm, align=center, font=\scriptsize, rounded corners=2pt, fill=armblue!5, minimum height=0.85cm},
      result/.style={rectangle, draw=paramgreen, very thick, text width=2.8cm, align=center, font=\scriptsize, rounded corners=2pt, fill=paramgreen!10, minimum height=0.85cm},
      arr/.style={->, >=stealth, thick, armblue},
    ]
      \node[box] (start) {$p > 1000$ features?};
      \node[result, right=of start] (filter) {Filter first (Variance + MI), then re-decide};
      \node[box, below=of start] (linear) {Using a linear model?};
      \node[result, right=of linear] (lasso) {Lasso / ElasticNet};
      \node[box, below=of linear] (tree) {Using a tree ensemble?};
      \node[result, right=of tree] (perm) {Permutation importance + RFECV};
      \node[box, below=of tree] (stable) {Need robust ``all-relevant''?};
      \node[result, right=of stable] (bor) {Boruta or stability selection};
      \node[result, below=of stable] (shap) {SHAP for production interpretability};

      \draw[arr] (start) -- node[above, font=\tiny]{yes} (filter);
      \draw[arr] (start) -- (linear);
      \draw[arr] (linear) -- node[above, font=\tiny]{yes} (lasso);
      \draw[arr] (linear) -- (tree);
      \draw[arr] (tree) -- node[above, font=\tiny]{yes} (perm);
      \draw[arr] (tree) -- (stable);
      \draw[arr] (stable) -- node[above, font=\tiny]{yes} (bor);
      \draw[arr] (stable) -- (shap);
    \end{tikzpicture}
  \end{center}
  \vfill
  \footnotesize Whatever method you use: do selection inside a CV loop, not before.
\end{frame}

% ================================================================
\section{Wrap-up}

\begin{frame}{Recap}
  \begin{itemize}\small
    \item \textbf{Why select.} Curse of dimensionality, multicollinearity, overfitting, compute, interpretability. Not always worth it - check vs ``keep all + regularize''.
    \item \textbf{Filter} - cheap, model-agnostic, blind to interactions. \texttt{SelectKBest} with \texttt{f\_regression} or \texttt{mutual\_info\_regression}.
    \item \textbf{Embedded} - selection as part of fitting. Lasso for linear, tree importances for trees. Watch for the correlated-features trap.
    \item \textbf{Permutation importance} - the reliable replacement for impurity importance. \texttt{SHAP} - the modern industry default.
    \item \textbf{Wrapper} - RFE-CV picks $k$ automatically. Boruta finds \emph{all relevant} features via shadow comparisons.
    \item \textbf{Stability selection} - bootstrap + Lasso, keep features that show up often.
    \item \textbf{Pitfall \#1:} feature selection on full data before splitting inflates CV scores. Always select inside the CV loop.
  \end{itemize}
\end{frame}

\begin{frame}{HW01h - practice}
  \footnotesize Estimated time: 90-120 minutes. Tasks 1-4 graded; bonuses optional.
  \begin{enumerate}\small
    \item On the Yerevan rent dataset (with engineered features from L01g): compute \texttt{f\_regression} scores. Plot scores and pick top 10 by univariate filter. Report Ridge CV MAE on the filtered set.
    \item Fit \texttt{LassoCV}. Plot the coefficient path (\texttt{lasso\_path}). Report which features survive at the CV-optimal $\lambda$ and compare to step 1.
    \item Fit a Random Forest. Compute \emph{both} \texttt{feature\_importances\_} (impurity) and \texttt{permutation\_importance} on a held-out set. Identify any feature where the two disagree by more than 2$\times$. Explain why.
    \item Run \texttt{RFECV} with Ridge. How does its chosen $k$ compare to the filter and Lasso choices?
    \item \emph{Bonus 1:} run stability selection (bootstrap + Lasso, 100 iterations). Plot selection frequency per feature. How does it compare to a single Lasso fit?
    \item \emph{Bonus 2:} install \texttt{boruta} (\texttt{pip install Boruta}) and run \texttt{BorutaPy}. Compare confirmed features to RFE-CV's selection.
    \item \emph{Bonus 3:} simulate selection bias: do feature selection on the \emph{full} dataset, then CV. Compare to doing selection inside the CV loop. Report the gap. \textbf{Most instructive question.}
  \end{enumerate}
\end{frame}

\end{document}

% =============================================================
% Provenance:
%   - Outline approved by user.
%   - Yerevan apartment running example continued from L01b-L01g.
%   - Callbacks: L01b (Ridge -> Lasso), L01g (filter/embedded/wrapper teaser),
%     math module 30 (curse of dim), L05b (nested CV for honest reporting).
%   - Predict-first frames per memory feedback_predict_first_slides.md.
% =============================================================
